---
title: "Adding a second classifier: global vs. local database reconciliation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-database workflow (global vs. local)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette is the optional add-on to `vignette("REGATTA-tutorial")`. Start
there for building and taxonomizing the regional checklist and the basic
single-classifier workflow; here we cover the extra step for when you have
**the same ASVs classified against two reference databases**.

## When you'd use this

**Why this matters:** a second, locally-curated database earns its keep two
ways — it catches species-level conflicts the global database would hide, and
it recovers regional species the global database has no reference for. But you
only get those gains if the two classifications are reconciled *honestly*
before the checklist step, rather than silently trusting whichever ran last.

A common eDNA setup classifies each ASV twice:

- against a **global** reference database (e.g. obitools + NCBI/EMBL), which has
  broad coverage but confidently names species from anywhere in the world, and
- against a **locally-curated** database (e.g. vsearch+SINTAX + a regional
  reference), which is sparser but regionally trustworthy.

`reconcile_global_local()` combines the two **before** the checklist step: per
ASV it keeps the call with the higher percent identity, and when the global
call wins but the local database disagrees, it downgrades to the lowest common
ancestor of the two — so a species-level conflict becomes an honest
genus-level call instead of a silent guess.

It's the **same Galapagos MiFish dataset** as the main tutorial: the identical
1,438 ASVs, now classified a second way.

## `run_regatta()`

**Why this matters:** adding a second database doesn't mean running a second
pipeline. `run_regatta()` runs `reconcile_global_local()` and then
`reconcile_checklist()` for you, in one call — so the two-database workflow is
still a single line.

**How:** pass both results as a named `list(global =, local =)`. Formats are
auto-detected (here the local side is a vsearch `lca` taxonomy file plus a
`--userout` file carrying the percent identities).

```{r run, eval = FALSE}
library(REGATTA)

res <- run_regatta(
  input = list(
    global = "MiFish_obitools.tab",                 # obitools (global NCBI DB)
    local  = c("MiFish_vsearch_lca.txt",            # vsearch LCA taxonomy
               "MiFish_vsearch_userout.txt")        # vsearch %ID
  ),
  checklist = cl$for_LCA,                            # from build_regional_checklist()
  sql_path  = "/path/to/accessionTaxa.sql"
  # out_dir = "regatta_out"                          # uncomment to also write the bundle
)
#> returns res$reconciled, res$post_checklist, res$summary.
#> with out_dir set, also writes reconcile_global_local/*, reconcile_checklist/*,
#> regatta_summary.csv, run_log.txt
```

On the full dataset, of the 1,438 ASVs, 892 were unassigned by both
classifiers and 546 were assigned by at least one. Stage 1 resolved those:

| | ASVs |
|---|---|
| best %ID: global won | 403 |
| best %ID: local won | 143 |
| global downgraded to global∩local LCA | 45 |

and Stage 2 (checklist) then gave 295 species / 231 genus / 19 family / 1 order
/ 892 off-target — versus the main tutorial's single-classifier obitools run
(287 species). Those extra species-level calls are the local database's
contribution.

## `reconcile_global_local()`

**Why this matters:** `run_regatta()` calls this for you, but running it
directly shows exactly how each ASV's two calls were reconciled — which won on
percent identity, and what REGATTA preferred. It runs live here, with no
internet and no database.

**How:** pass the two classifier tables. The demo uses **real slices of the
Galapagos MiFish dataset** bundled with the package — 12 real ASVs classified
both ways, chosen to show each decision (you would normally pass the whole
tables).

```{r demo}
library(REGATTA)

global <- read.csv(system.file("extdata", "mifish_obitools_example.csv",
                               package = "REGATTA"), stringsAsFactors = FALSE)
local  <- read.csv(system.file("extdata", "mifish_vsearch_example.csv",
                               package = "REGATTA"), stringsAsFactors = FALSE)

rec <- reconcile_global_local(global, local, output_dir = NULL)
```

`$tracking` shows the per-ASV decision — both classifiers' calls side by side,
which won on percent identity, and the database REGATTA preferred:

```{r demo-tracking}
rec$tracking[, c("ASV_id", "species_global", "pct_id_global",
                 "species_local", "pct_id_local",
                 "best_pctid_winner", "preferred_database",
                 "preferred_scientific_name")]
```

There are **four** per-ASV outcomes, two genera each here:

* **Global only** (the local database made no call → keep global): the six
  checklist-demo genera from the main tutorial — *Abudefduf saxatilis* /
  *taurus*, *Prionurus laticlavius* / *biafraensis*, *Archosargus rhomboidalis*,
  *Conodon nobilis*. With nothing to reconcile, they pass through unchanged.
* **Both agree → kept** — *Lutjanus kasmira* and *Caranx melampygus*: both
  databases name the same species (with `Local_advantage = TRUE`, the default,
  the tie goes to local). The species is kept.
* **Disagree, global %ID wins → LCA downgrade** (`global_lca_to_local`) —
  *Epinephelus bontoides* (global) vs. *E. merra* (local), and *Mugil curema*
  (global) vs. *M. thoburni* (local): the global call has the higher identity,
  but because the two conflict at species, REGATTA reports the lowest common
  ancestor — genus ***Epinephelus*** and genus ***Mugil***.
* **Disagree, local %ID wins → local kept** — *Etrumeus teres* (local, 99.4%)
  over *E. micropus* (global, 98.9%), and *Cubiceps pauciradiatus* (local) over
  *C. squamiceps* (global): here the locally-curated database has the better
  match, so its call is trusted outright.

Feed the reconciled result straight into the checklist step, exactly as in the
single-classifier workflow:

```{r demo-checklist}
checklist <- readRDS(system.file("extdata", "galapagos_fish_checklist.rds",
                                 package = "REGATTA"))
post <- reconcile_checklist(rec$result, checklist, output_dir = NULL)
```

## Same 12 ASVs, both methods, side by side

**Why this matters:** whether a second database is worth it comes down to what
actually changes versus the single-classifier run — and the changes go both
ways: some calls become *less* specific (a conflict caught), some become *more*
specific (a regional species recovered).

**How:** run the main-tutorial method (obitools straight to the checklist) on
the same inputs and compare the final call:

```{r compare}
single <- reconcile_checklist(global, checklist, output_dir = NULL)

lowest <- function(r) {            # the most specific non-NA rank per ASV
  ranks <- c("domain","phylum","class","order","family","genus","species")
  apply(r$result[, ranks], 1, function(x) {nz <- which(!is.na(x)); if (length(nz)) x[max(nz)] else NA})
}
data.frame(
  obitools_call = global$species,
  single_db     = lowest(single),
  two_db        = lowest(post)
)
```

Nine of the twelve land in the same place — but **three change**, and each in a
different, instructive way:

* **Mugil curema** — single-DB keeps the species (*M. curema* is on the
  checklist); two-DB drops to genus ***Mugil*** because the local database's
  *M. thoburni* contradicts it. Two databases caught a conflict one couldn't.
* **Etrumeus micropus** — single-DB keeps the global species; two-DB replaces it
  with ***Etrumeus teres***, the local database's higher-identity (and
  regionally-curated) call.
* **Cubiceps squamiceps** — single-DB has to downgrade to genus *Cubiceps*
  (*squamiceps* isn't on the checklist); two-DB keeps the **species**
  ***Cubiceps pauciradiatus***, because the local database's call *is* on the
  checklist. Here the second database makes the result **more** specific, not
  less.

## Inspecting and tuning

**Why this matters:** the reconciliation is fully auditable and adjustable — you
can see why each ASV got its call, and change how ties and percent-identity
scales are handled to fit your databases.

**How:**

- The full Stage 1 audit (`reconcile_global_local/` on disk, or `rec$tracking`)
  carries every input column from both classifiers suffixed `_global` /
  `_local`, plus `preferred_pctid` and the reconciled `preferred_<rank>`
  lineage — so you can see exactly why each ASV got its call.
- `Local_advantage = FALSE` makes ties go to the global database instead.
- Percent-identity scales are auto-detected per input (obitools' 0–1
  `BEST_IDENTITY` is rescaled to 0–100 to match vsearch); override with
  `global_pct_id_scale` / `local_pct_id_scale` if needed.

See `?reconcile_global_local` for the full argument list, and
`vignette("REGATTA-tutorial")` for the checklist-building steps this workflow
assumes.
