---
title: "REGATTA tutorial: reconciling eDNA assignments against a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

REGATTA reconciles eDNA taxonomic assignments with the species actually known
from your study region. When a classifier assigns a read to a species that is
taxonomically plausible but geographically unlikely, REGATTA walks that
assignment up the taxonomic tree only as far as the regional checklist requires:
species-level calls are kept where the species is regionally supported, and
unsupported calls are downgraded to the lowest supported rank rather than
discarded.

This tutorial runs the whole workflow end to end on a Galapagos reef-fish
example — build the regional checklist (**Module 1**), then reconcile a
classifier's output against it (**Module 2**). Run the chunks in order; each one
works as written (with an internet connection).

You provide three things: the taxonomic group your primer targets (`"fish"`
here), a regional polygon as a WKT string, and your classifier's output table.
Reconciling the same reads against a second (global + local) database is
**Module 3**, covered in `vignette("REGATTA-two-database")`; optional pre-checks,
GBIF setup, tuning, and other classifier formats are in
`vignette("REGATTA-advanced")`.

## Setup

Load the package and choose where the local NCBI taxonomy snapshot lives. REGATTA
uses this snapshot (built by `taxonomizr`) to resolve names and taxIDs to full
lineages. By default we keep it in the working directory; it is built on first
use — a few hundred MB, a few minutes — and reused afterwards. The first checklist
build below will offer to create it.

```{r setup-run}
library(REGATTA)

sql_path <- "accessionTaxa.sql"   # stored in your working directory; edit to taste
```

## Module 1: build the regional checklist

`build_regional_checklist()` downloads a regional species list (OBIS by default),
resolves it to NCBI taxonomy, and returns the pieces you need — writing them to
`output_dir` as well. Draw your own study-area polygon at
[wktmap.com](https://wktmap.com/) and paste the WKT string; the one below is a
box around the Galapagos.

```{r build, eval = FALSE}
poly <- "POLYGON ((-92 2, -89 2, -89 -2, -92 -2, -92 2))"

cl <- build_regional_checklist(
  region        = "galapagos",
  label         = "fish",
  taxa          = "fish",
  regional_poly = poly,
  OBIS          = TRUE,
  marine        = TRUE,
  terrestrial   = FALSE,
  output_dir    = "regatta_checklist",
  sql_path      = sql_path
)
```

`cl` holds the checklist in a few forms:

| Element | Contents |
|---|---|
| `cl$for_LCA` | The taxID column plus the seven ranks — pass this to `run_regatta()`. |
| `cl$for_making_localdb` | Unique species binomials, also written one per line for reference-database builders such as CRABS. |
| `cl$checklist_summary` | The full taxonomized checklist, showing how each name resolved. |
| `cl$methods` | A ready-to-adapt methods and provenance record, with citations. |

## Module 2: reconcile a classifier against the checklist

`run_regatta()` reads your classifier output, resolves lineages, reconciles each
call against the checklist, and returns the final tables (writing them to
`out_dir`). Point `input` at your own results file; here we use the obitools
example bundled with the package so the chunk runs as written.

```{r run, eval = FALSE}
input_tab <- system.file("extdata", "mifish_obitools_example.csv",
                         package = "REGATTA")

res <- run_regatta(
  input     = input_tab,
  checklist = cl$for_LCA,
  out_dir   = "regatta_out",
  region    = "galapagos",
  label     = "fish"
)

res$taxonomy_table
res$summary
```

## Interpreting the outputs

`run_regatta()` returns three things — also written to a dated
`<region>_<label>_<date>/` subfolder of `out_dir` — and every REGATTA step that
produces a taxonomy table returns the same shape, so results drop straight into
phyloseq or MetabaR and stay auditable:

| Element | Contents |
|---|---|
| `res$taxonomy_table` | The reconciled exchange-format table: `ASV_id`, `pct_id` when available (normalized to 0–100), then the seven ranks. For a phyloseq `tax_table()`, drop `pct_id` first. |
| `res$tracking` | Per-ASV audit: before/after taxonomy, the rank each call matched, the reconciliation reason, and passthrough columns. |
| `res$summary` | Per-stage counts — kept, downgraded, dropped, and the per-rank-pair downgrade breakdown. |

The tracking table is where you check why a call changed. When the checklist
carries a target group, `regatta_reason` distinguishes outcomes such as `kept`,
`non-local (geographic)`, and `off-target (taxonomic)`.

That is the whole workflow. For a second classifier (Module 3) see
`vignette("REGATTA-two-database")`; for pre-checking taxon names, GBIF setup,
tuning, inspecting the checklist step ASV by ASV, and other classifier formats,
see `vignette("REGATTA-advanced")`.
