---
title: "REGATTA tutorial: correcting off-target eDNA assignments with a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What REGATTA does, and what you need

Global reference databases (NCBI, EMBL, …) confidently assign eDNA reads to
species that don't live in your study area, because the closest match in the
database is a relative from somewhere else in the world. REGATTA corrects this
**without hand-curating a local reference database and without throwing away
taxonomic specificity**: it builds a regional species checklist from public
biodiversity records, then walks each assignment up the taxonomic tree only as
far as the regional checklist forces.

To run the whole pipeline you supply **three things**:

1. **The taxonomic group(s)** your primer targets, as taxon names.
2. **A regional polygon** — a WKT `POLYGON((...))` string. Draw it at
   [wktmap.com](https://wktmap.com) and copy the generated text.
3. **A classifier results file** — whatever your metabarcoding taxonomy-
   assignment step produced (obitools `.tab`, vsearch, etc.). REGATTA
   auto-detects the format.

This tutorial walks the full path on the package's worked example: **Galapagos
reef fish from the MiFish primer, classified with obitools against the global
NCBI database.** Substitute your own group, polygon, and results file — the
function names and shapes don't change.

> **The pipeline is not fish-, marine-, or Galapagos-specific.** The only
> per-region work is building and taxonomizing the checklist; every downstream
> step is taxonomy- and geography-agnostic.

---

## Naming your taxa (you don't need to know the rank quirks)

MiFish is a fish/vertebrate primer; for it you want **ray-finned fishes plus
elasmobranchs (sharks and rays)**. The historical headache is that the public
databases disagree on names and ranks — WoRMS calls ray-finned fish
`Actinopteri`, NCBI uses `Actinopterygii`, much of the literature says
`Teleostei` or `Osteichthyes`, and GBIF's backbone skips the fish class
entirely. REGATTA hides all of that behind `resolve_taxa()`. There's even a
one-word shorthand — **`"fish"`** expands to ray-finned fishes + elasmobranchs:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("fish")
#> input  alias_used      aphia_id  valid_name      worms_status  gbif_usable
#> fish   Actinopterygii  10194     Actinopterygii  ok            FALSE
#> fish   Elasmobranchii  10193     Elasmobranchii  ok            TRUE
#> fish   Myxini          10189     Myxini          ok            TRUE
#> fish   Petromyzonti    843662    Petromyzonti    ok            TRUE
```

`resolve_taxa()` (1) disambiguates names against WoRMS *by kingdom* — so an
ambiguous name like `"Vertebrata"` resolves to the vertebrate subphylum, not
the red-algae genus of the same name — and (2) reports whether GBIF's backbone
has a usable key. `gbif_usable = FALSE` for ray-finned fish is expected: GBIF
has no class node for them, so `GBIF_download()` automatically descends to the
**order** rank GBIF *does* populate (and unions with direct keys for
sharks/mammals/birds). You never have to know that. `"fish"`, `Teleostei`,
`Actinopteri`, `Actinopterygii`, and `Osteichthyes` all behave identically;
built-in shorthands also include `"vertebrates"` (→ Vertebrata). (`"fish"`
covers ray-finned fishes, sharks & rays, hagfishes, and lampreys. The whole
GBIF search is a single `occ_download`, so by default it goes for completeness
— ~98%; pass `gbif_fill_families = FALSE` to `GBIF_download()` for a faster,
order-only ~97% preparation.)

---

## Step 1 — Build the regional checklist (once per region × group)

Pull species lists from OBIS (primary) and GBIF (supplement) inside your
polygon, then stack them. These calls need internet and a GBIF account (see the
README "Setup" section), so they're shown but not run here.

```{r build, eval = FALSE}
poly <- "POLYGON ((-117.421875 31.952162, -91.933594 -6.315299, -81.386719 -6.315299, -76.113281 7.710992, -82.089844 8.581021, -87.011719 13.581921, -104.238281 20.303418, -112.5 32.249974, -117.421875 31.952162))"  # E. Tropical Pacific incl. Galapagos

# OBIS — primary. Queries by AphiaID under the hood, so any fish synonym
# works; "fish" expands to ray-finned fishes + elasmobranchs.
OBIS_download(
  obis_taxa       = "fish",
  regional_poly   = poly,
  obis_outputname = "OBIS_galapagos_fish"
)

# GBIF — supplement. One occ_download: order keys for ray-finned fish, plus a
# single class key each for sharks/rays, hagfishes, and lampreys, plus
# family gap-fill (default) for ~98% coverage.
GBIF_download(
  obis_taxa       = "fish",
  regional_poly   = poly,
  gbif_outputname = "GBIF_galapagos_fish"
)

# Stack + dedupe GBIF and OBIS into one regional list, optionally folding in
# a local checklist CSV (needs Genus + Species columns) from anywhere on
# disk -- no need to copy it into the project first.
build_regional_checklist(
  region          = "galapagos",
  taxa            = "fish",
  comb_inputnames = c("OBIS_galapagos_fish", "GBIF_galapagos_fish"),
  local_csvs      = "2016Aug24_Tirado-Sanchez_et_al_Galapagos_Pisces_Checklist.csv"
)
# Writes two files under local_database_checklist/:
#   comprehensive_galapagos_fish_list_for_making_localdb.txt  (species only)
#   comprehensive_galapagos_fish_list_for_LCA.txt             (+ retained genera)
```

## Step 2 — Taxonomize the checklist (once per region × group)

Resolve every species name to a full 7-rank NCBI lineage. This builds (first
time, ~15 min) or reuses a local `taxonomizr` database. The lookup is
synonym-aware, so older names are updated to current canonical NCBI taxonomy
automatically.

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "local_database_checklist/comprehensive_galapagos_fish_list_for_LCA.txt",
  sql_path = "/path/to/accessionTaxa.sql"
)
saveRDS(checklist, "local_database_checklist/galapagos_fish_checklist.rds")
```

## Step 3 — Reconcile your obitools output (the one-call path)

`run_regatta()` auto-detects the obitools `.tab` (it has a `TAXID` column),
resolves the taxIDs to lineages, runs the checklist LCA, and writes the output
bundle.

```{r run, eval = FALSE}
run_regatta(
  input     = "MiFish_obitools.tab",                   # obitools .tab (global DB)
  checklist = "local_database_checklist/galapagos_fish_checklist.rds",
  sql_path  = "/path/to/accessionTaxa.sql"
)
#> -> regatta_out/{ reconcile_checklist/*, regatta_summary.csv, run_log.txt }
```

On this dataset, of the 542 obitools-assigned ASVs (out of 1,438 total):

| match rank | ASVs |
|---|---|
| species | 287 |
| genus   | 229 |
| family  | 16 |
| order   | 1 |
| class   | 9 |

The 229 genus-level and 9 class-level results are the off-target calls REGATTA
exists to catch — globally-confident species assignments to fish that don't
occur in the Galapagos, walked up to the lowest rank the regional checklist
supports.

> **Classified the same ASVs against a second, locally-curated database too?**
> REGATTA can reconcile the two classifier runs against each other before the
> checklist step — see `vignette("REGATTA-two-database")`.

---

## A runnable demonstration (no internet, no database)

The core `reconcile_checklist()` step is pure base R. It runs live here against
two files bundled with the package — both **real slices of the Galapagos MiFish
dataset**: the taxonomized regional checklist, and **12 real obitools
assignments (10 genera)** chosen to show every outcome. The two-database
vignette puts these *same 12 ASVs* through the optional global-vs-local step, so
you can compare the methods directly. You would normally pass the *whole*
classifier table, not a hand-picked slice.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file(
  "extdata", "galapagos_fish_checklist.rds", package = "REGATTA"))

# 12 real obitools-assigned ASVs from the worked dataset (ASV_id + 7 ranks + pct_id).
tax <- read.csv(system.file(
  "extdata", "mifish_obitools_example.csv", package = "REGATTA"),
  stringsAsFactors = FALSE)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)
```

`$tracking` records *why* each row changed — the input species, the percent
identity, and the rank at which it matched the regional checklist:

```{r demo-tracking}
res$tracking[, c("before_species", "pct_id", "regatta_match_rank",
                 "after_genus", "after_family")]
```

The checklist step has **three outcomes**, and all three appear here:

* **Kept at species** (the call is on the checklist) — *Abudefduf saxatilis*,
  *Prionurus laticlavius*, *Lutjanus kasmira*, *Caranx melampygus*,
  *Mugil curema*, *Etrumeus micropus*.
* **Downgraded to genus** (species off the checklist, genus on) —
  *Abudefduf taurus* → *Abudefduf*, *Prionurus biafraensis* → *Prionurus*,
  *Epinephelus bontoides* → *Epinephelus*, *Cubiceps squamiceps* → *Cubiceps*.
* **Downgraded to family** (species *and* genus off the checklist) —
  *Archosargus rhomboidalis* → **Sparidae**, *Conodon nobilis* → **Haemulidae**.

The two *within-genus splits* make the point sharpest. *Abudefduf saxatilis*
and *A. taurus* **both matched the global database at 100% identity**, yet only
*saxatilis* is on the Galapagos checklist, so *taurus* drops to genus — a
percent-identity cutoff could never separate them. Likewise *Prionurus
laticlavius* (the Galapagos sawtail surgeonfish) stays at species while
*P. biafraensis* — a **West African** congener confidently called at 98.8% —
drops to genus. The global database simply has no reference for some Galapagos
species, so it names the nearest relative it has; REGATTA keeps the reliable
genus and drops the unreliable species.

`$stats` gives the match-rank distribution for this slice:

```{r demo-stats}
res$stats
```

> Six of these ASVs were *kept at species* here — but three of those calls
> change once a second, locally-curated classifier weighs in.
> `vignette("REGATTA-two-database")` runs the same 12 ASVs through that step and
> compares the two methods side by side.

---

## Interpreting the outputs

Every REGATTA function that produces a taxonomy table returns the same three
elements (and, when writing to disk, the same three CSVs per stage):

| Element | Contents |
|---|---|
| `$result`   | The **exchange format**: exactly the id column + 7 lowercase ranks. Drop straight into a phyloseq `tax_table()` or a MetabaR MOTU table. |
| `$tracking` | Per-ASV audit: before/after at each rank, `regatta_match_rank`, and your passthrough columns. |
| `$stats`    | Step-level counts — the match-rank distribution above. |

## Where to go next

* **A different results format?** `run_regatta()` auto-detects vsearch (LCA +
  `--userout`), and Kraken2/BestTaxon-style name tables. For a table whose
  per-ASV id column isn't `ASV_id` (e.g. a sequence `Hash`), pass
  `id_col = "Hash"`; a long per-sample table is collapsed to one row per ASV
  automatically. See `?run_regatta`.
* **Two reference databases?** If you classified the same ASVs against both a
  global and a locally-curated database, REGATTA can reconcile the two
  classifier runs against each other before the checklist step. See
  `vignette("REGATTA-two-database")` and `?reconcile_global_local`.
* **One pass per primer.** Run the pipeline once per taxonomic group your
  primers target. Mixing groups into one megalist lets off-target
  amplifications slip through.
* **Pre-check your taxon names** with `resolve_taxa()` before a long download,
  and see the README troubleshooting section for GBIF/OBIS specifics.
