---
title: "REGATTA tutorial: correcting off-target eDNA assignments with a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What REGATTA does, and what you need

Global reference databases (NCBI, EMBL, …) confidently assign eDNA reads to
species that don't live in your study area, because the closest match in the
database is a relative from somewhere else in the world. REGATTA corrects this
**without hand-curating a local reference database and without throwing away
taxonomic specificity**: it builds a regional species checklist from public
biodiversity records, then walks each assignment up the taxonomic tree only as
far as the regional checklist forces.

To run the whole pipeline you supply **three things**:

1. **The taxonomic group(s)** your primer targets, as taxon names.
2. **A regional polygon** — a WKT `POLYGON((...))` string. Draw it at
   [wktmap.com](https://wktmap.com) and copy the generated text.
3. **A classifier results file** — whatever your metabarcoding taxonomy-
   assignment step produced (obitools `.tab`, vsearch, etc.). REGATTA
   auto-detects the format. If you classified the same ASVs against **two**
   databases (a global reference DB and a locally-curated one), you can supply
   both and REGATTA will reconcile them first.

This tutorial walks the full path on the package's worked example: **Galapagos
reef fish from the MiFish primer, classified two ways** — with obitools against
the global NCBI database *and* with vsearch+SINTAX against a locally-curated
reference DB. REGATTA reconciles the two classifier runs against each other,
then against the regional checklist. Substitute your own group, polygon, and
results file(s) — the function names and shapes don't change.

> **The pipeline is not fish-, marine-, or Galapagos-specific.** The only
> per-region work is building and taxonomizing the checklist; every downstream
> step is taxonomy- and geography-agnostic.

---

## Naming your taxa (you don't need to know the rank quirks)

MiFish is a fish/vertebrate primer; for it you want **ray-finned fishes plus
elasmobranchs (sharks and rays)**. The historical headache is that the public
databases disagree on names and ranks — WoRMS calls ray-finned fish
`Actinopteri`, NCBI uses `Actinopterygii`, much of the literature says
`Teleostei` or `Osteichthyes`, and GBIF's backbone skips the fish class
entirely. REGATTA hides all of that behind `resolve_taxa()`. There's even a
one-word shorthand — **`"fish"`** expands to ray-finned fishes + elasmobranchs:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("fish")
#> input  alias_used      aphia_id  valid_name      worms_status  gbif_usable
#> fish   Actinopterygii  10194     Actinopterygii  ok            FALSE
#> fish   Elasmobranchii  10193     Elasmobranchii  ok            TRUE
#> fish   Myxini          10189     Myxini          ok            TRUE
#> fish   Petromyzonti    843662    Petromyzonti    ok            TRUE
```

`resolve_taxa()` (1) disambiguates names against WoRMS *by kingdom* — so an
ambiguous name like `"Vertebrata"` resolves to the vertebrate subphylum, not
the red-algae genus of the same name — and (2) reports whether GBIF's backbone
has a usable key. `gbif_usable = FALSE` for ray-finned fish is expected: GBIF
has no class node for them, so `GBIF_download()` automatically descends to the
**order** rank GBIF *does* populate (and unions with direct keys for
sharks/mammals/birds). You never have to know that. `"fish"`, `Teleostei`,
`Actinopteri`, `Actinopterygii`, and `Osteichthyes` all behave identically;
built-in shorthands also include `"vertebrates"` (→ Vertebrata). (`"fish"`
covers ray-finned fishes, sharks & rays, hagfishes, and lampreys. The whole
GBIF search is a single `occ_download`, so by default it goes for completeness
— ~98%; pass `gbif_fill_families = FALSE` to `GBIF_download()` for a faster,
order-only ~97% preparation.)

---

## Step 1 — Build the regional checklist (once per region × group)

Pull species lists from OBIS (primary) and GBIF (supplement) inside your
polygon, then stack them. These calls need internet and a GBIF account (see the
README "Setup" section), so they're shown but not run here.

```{r build, eval = FALSE}
poly <- "POLYGON ((-117.421875 31.952162, -91.933594 -6.315299, -81.386719 -6.315299, -76.113281 7.710992, -82.089844 8.581021, -87.011719 13.581921, -104.238281 20.303418, -112.5 32.249974, -117.421875 31.952162))"  # E. Tropical Pacific incl. Galapagos

# OBIS — primary. Queries by AphiaID under the hood, so any fish synonym
# works; "fish" expands to ray-finned fishes + elasmobranchs.
OBIS_download(
  obis_taxa       = "fish",
  regional_poly   = poly,
  obis_outputname = "OBIS_galapagos_fish"
)

# GBIF — supplement. One occ_download: order keys for ray-finned fish, plus a
# single class key each for sharks/rays, hagfishes, and lampreys, plus
# family gap-fill (default) for ~98% coverage.
GBIF_download(
  obis_taxa       = "fish",
  regional_poly   = poly,
  gbif_outputname = "GBIF_galapagos_fish"
)

# Optional: add a local checklist CSV (needs Genus + Species columns).
Local_csv_download(
  loc_csvs       = "2016Aug24_Tirado-Sanchez_et_al_Galapagos_Pisces_Checklist.csv",
  loc_outputname = "Local_galapagos_fish"
)

# Stack + dedupe into one regional list.
build_regional_checklist(
  comb_inputnames = c("OBIS_galapagos_fish", "GBIF_galapagos_fish",
                      "Local_galapagos_fish"),
  comb_outputname = "comprehensive_galapagos_fish_list"
)
```

## Step 2 — Taxonomize the checklist (once per region × group)

Resolve every species name to a full 7-rank NCBI lineage. This builds (first
time, ~15 min) or reuses a local `taxonomizr` database. The lookup is
synonym-aware, so older names are updated to current canonical NCBI taxonomy
automatically.

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "custom_db/comprehensive_galapagos_fish_list.txt",
  sql_path = "/path/to/accessionTaxa.sql"
)
saveRDS(checklist, "custom_db/galapagos_fish_checklist.rds")
```

## Step 3 — Reconcile two classifier runs, then the checklist

Here the same 1,438 MiFish ASVs were classified twice: with **obitools against
the global NCBI database** (a `.tab` file with a `TAXID` column) and with
**vsearch+SINTAX against a locally-curated reference DB** (a `lca` taxonomy
file plus a `--userout` file carrying percent identities). Give both to
`run_regatta()` as a named `list(global=, local=)` and it does two things in
one call: reconciles the two classifier runs against each other
([reconcile_global_local()]), then reconciles the result against the regional
checklist ([reconcile_checklist()]). Formats are auto-detected.

```{r run, eval = FALSE}
run_regatta(
  input = list(
    global = "MiFish_obitools.tab",                       # obitools (global DB)
    local  = c("MiFish_vsearch_lca.txt",                  # vsearch LCA taxonomy
               "MiFish_vsearch_userout.txt")              # vsearch %ID
  ),
  checklist = "custom_db/galapagos_fish_checklist.rds",
  sql_path  = "/path/to/accessionTaxa.sql"
)
#> -> regatta_out/{ reconcile_global_local/*, reconcile_checklist/*,
#>                  regatta_summary.csv, run_log.txt }
```

**Stage 1 — global vs. local (`reconcile_global_local`).** Of 1,438 ASVs, 892
were unassigned by both classifiers; 546 were assigned by at least one. Per
ASV, the database with the higher percent identity wins, and when global wins
but local also has a call, the row is downgraded to the LCA of the two:

| | ASVs |
|---|---|
| best %ID: global won | 403 |
| best %ID: local won | 143 |
| global downgraded to global∩local LCA | 45 |

**Stage 2 — reconcile against the checklist (`reconcile_checklist`).** The 546
reconciled assignments are then walked against the Galapagos fish checklist:

| match rank | ASVs |
|---|---|
| species | 295 |
| genus   | 231 |
| family  | 19 |
| order   | 1 |
| off-target (no match) | 892 |

The 231 genus-level and 19 family-level results are the off-target calls
REGATTA exists to catch — confident species assignments to fish that don't
occur in the Galapagos, walked up to the lowest rank the regional checklist
supports.

> **Only have one classifier run?** Pass a single file (or a single vsearch
> `lca` + `userout` pair) as `input` instead of the `list(global=, local=)`;
> `run_regatta()` skips Stage 1 and goes straight to the checklist step.

---

## A runnable demonstration (no internet, no database)

The core `reconcile_checklist()` step is pure base R. Here it runs live against
the **real** Galapagos fish checklist bundled with the package, on a small
illustrative table chosen to show each kind of downgrade — including the
elasmobranch (shark) side of a MiFish run.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file(
  "extdata", "galapagos_fish_checklist.rds", package = "REGATTA"))

# A small classifier-output table (ASV_id + the 7 lowercase rank columns).
tax <- data.frame(
  ASV_id  = paste0("ASV_", 1:6),
  domain  = "Eukaryota",
  phylum  = c("Chordata", "Chordata", "Chordata", "Chordata", "Chordata", "Echinodermata"),
  class   = c("Actinopteri", "Actinopteri", "Actinopteri", "Actinopteri", "Elasmobranchii", "Echinoidea"),
  order   = c("Perciformes", "Perciformes", "Scorpaeniformes", "Mulliformes", "Carcharhiniformes", "Camarodonta"),
  family  = c("Lutjanidae", "Lutjanidae", "Sebastidae", "Mullidae", "Carcharhinidae", "Strongylocentrotidae"),
  genus   = c("Lutjanus", "Lutjanus", "Sebastes", "Mulloidichthys", "Triaenodon", "Strongylocentrotus"),
  species = c("Lutjanus argentiventris", "Lutjanus campechanus", "Sebastes goodei",
              "Mulloidichthys dentatus", "Triaenodon obesus", "Strongylocentrotus purpuratus"),
  stringsAsFactors = FALSE
)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)
```

`$result` is the corrected taxonomy table — the strict 8-column REGATTA
exchange format (id + 7 ranks), with ranks below each match set to `NA`:

```{r demo-result}
res$result
```

`$tracking` records *why* each row changed — the rank at which it matched the
regional checklist:

```{r demo-tracking}
res$tracking[, c("ASV_id", "before_species", "regatta_match_rank",
                 "after_genus", "after_class")]
```

Reading it row by row:

* **Lutjanus argentiventris** (yellow snapper), **Mulloidichthys dentatus**
  (Mexican goatfish), **Triaenodon obesus** (whitetip reef shark) — all on the
  Galapagos checklist, kept at **species**. (The shark confirms the
  elasmobranch side of a MiFish run is handled, not just bony fish.)
* **Lutjanus campechanus** — the northern red snapper, an *Atlantic* species
  that can't be in a Galapagos sample. The species isn't on the checklist but
  the genus is, so it downgrades to genus ***Lutjanus*** instead of asserting a
  wrong species.
* **Sebastes goodei** — a temperate Pacific rockfish; the whole genus is absent
  from the tropical Galapagos, so it walks up to class **Actinopteri**.
* **Strongylocentrotus purpuratus** — a purple urchin (Echinodermata, an
  off-target invertebrate amplification); nothing below *Eukaryota* is on a
  fish checklist, so it walks all the way up to **domain**.

`$stats` gives the match-rank distribution for the whole table:

```{r demo-stats}
res$stats
```

---

## Interpreting the outputs

Every REGATTA function that produces a taxonomy table returns the same three
elements (and, when writing to disk, the same three CSVs per stage):

| Element | Contents |
|---|---|
| `$result`   | The **exchange format**: exactly the id column + 7 lowercase ranks. Drop straight into a phyloseq `tax_table()` or a MetabaR MOTU table. |
| `$tracking` | Per-ASV audit: before/after at each rank, `regatta_match_rank`, and your passthrough columns. |
| `$stats`    | Step-level counts — the match-rank distribution above. |

## Where to go next

* **A different results format?** `run_regatta()` auto-detects vsearch (LCA +
  `--userout`), and Kraken2/BestTaxon-style name tables. For a table whose
  per-ASV id column isn't `ASV_id` (e.g. a sequence `Hash`), pass
  `id_col = "Hash"`; a long per-sample table is collapsed to one row per ASV
  automatically. See `?run_regatta`.
* **Inspect the global-vs-local decisions.** The Stage 1 output
  (`reconcile_global_local/`) carries a per-ASV audit — `preferred_database`
  (`global` / `local` / `global_lca_to_local`), `preferred_pctid`, and both
  classifiers' lineages side by side — so you can see exactly why each ASV got
  the call it did. See `?reconcile_global_local`.
* **One pass per primer.** Run the pipeline once per taxonomic group your
  primers target. Mixing groups into one megalist lets off-target
  amplifications slip through.
* **Pre-check your taxon names** with `resolve_taxa()` before a long download,
  and see the README troubleshooting section for GBIF/OBIS specifics.
