---
title: "Adding a second classifier: global vs. local database reconciliation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-database workflow (global vs. local)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette covers Module 3 of REGATTA: reconciling the same ASVs classified
against two reference databases, a broad global one and a locally curated one,
to keep the benefits of both, then reconciling the result against the regional
checklist.

It builds on `vignette("REGATTA-tutorial")`, which covers Module 1 (building and
taxonomizing the regional checklist) and Module 2 (reconciling one classifier
against it). Start there first. Module 3 inserts one extra step,
`reconcile_global_local()`, ahead of the Module 2 checklist reconciliation, and
`run_regatta()` runs both in a single call.

## When you'd use this

A second, locally curated database earns its keep two ways. It catches
species-level conflicts the global database would hide, and it recovers regional
species the global database has no reference for. You only get those gains if the
two classifications are reconciled before the checklist step, rather than
trusting whichever ran last.

A common eDNA setup classifies each ASV twice:

- against a global reference database (here, obitools against sequences from
  EMBL), which has broad coverage but confidently names species from anywhere in
  the world, and
- against a locally curated database (here, vsearch against a regional reference
  curated from NCBI sequences), which is sparser but regionally trustworthy.

`reconcile_global_local()` combines the two before the checklist step. For each
ASV, it keeps the call with the higher percent identity. When the global call
wins but the local database disagrees, it downgrades to the lowest common
ancestor of the two, so a species-level conflict becomes a genus-level call
instead of a silent guess.

This vignette uses the same Galapagos MiFish dataset as the main tutorial: the
identical 1,438 ASVs, now classified a second way.

## `run_regatta()`

Use `run_regatta()` to run the two-database workflow in one call. It runs
`reconcile_global_local()` and then `reconcile_checklist()` for you, so adding a
second database does not mean running a second pipeline.

Pass both results as a named `list(global =, local =)`. Formats are
auto-detected. Here the local side is a vsearch `lca` taxonomy file plus a
`--userout` file carrying the percent identities.

```{r run, eval = FALSE}
library(REGATTA)

res <- run_regatta(
  input = list(
    global = "MiFish_obitools.tab",                 # obitools (global EMBL DB)
    local  = c("MiFish_vsearch_lca.txt",            # vsearch LCA taxonomy
               "MiFish_vsearch_userout.txt")        # vsearch %ID
  ),
  checklist = cl$for_LCA,                            # from build_regional_checklist()
  out_dir   = "regatta_out",                         # REQUIRED: dated run subfolder
  region    = "galapagos", label = "fish"            # name the run subfolder
  # sql_path defaults to the cache; pass a path to override.
)
#> returns res$taxonomy_table (final reconciled table), res$tracking (per-ASV
#> audit, both stages combined), res$summary, plus res$reconciled / res$post_checklist;
#> writes reconcile_global_local/*, reconcile_checklist/*, regatta_summary.csv,
#> run_log.txt into regatta_out/galapagos_fish_<date>/
```

On the full dataset, of the 1,438 ASVs, 892 were unassigned by both classifiers
and 546 were assigned by at least one. Stage 1 resolved those:

| | ASVs |
|---|---|
| best %ID: global won | 403 |
| best %ID: local won | 143 |
| global downgraded to global∩local LCA | 45 |

Stage 2 (the checklist) then gave 295 species, 231 genus, 19 family, and 1 order
among the 546 assigned. The checklist dropped none, because every reconciled call
had a regional record; the other 892 ASVs were simply unassigned by both
classifiers. Compared with the main tutorial's single-classifier obitools run
(287 species), those extra species-level calls are the local database's
contribution.

## `reconcile_global_local()`

`reconcile_global_local()` is the Stage 1 step that `run_regatta()` calls for
you. Running it directly shows how each ASV's two calls were reconciled: which
won on percent identity, and what REGATTA preferred. It runs with no internet and
no taxonomy database.

The demo uses real slices of the Galapagos MiFish dataset bundled with the
package: 12 real ASVs classified both ways, chosen to show each decision. You
would normally pass the whole tables.

```{r demo}
library(REGATTA)

global <- read.csv(system.file("extdata", "mifish_obitools_example.csv",
                               package = "REGATTA"), stringsAsFactors = FALSE)
local  <- read.csv(system.file("extdata", "mifish_vsearch_example.csv",
                               package = "REGATTA"), stringsAsFactors = FALSE)

rec <- reconcile_global_local(global, local, output_dir = NULL)
```

`$tracking` shows the per-ASV decision with each field grouped global / local /
regatta (the reconciled value), so you can read the two classifiers' calls and
REGATTA's pick side by side:

```{r demo-tracking}
rec$tracking[, c("ASV_id",
                 "species_global", "species_local", "species_regatta",
                 "pct_id_global", "pct_id_local", "pct_id_regatta",
                 "best_pctid_winner", "preferred_database")]
```

There are four per-ASV outcomes, with two genera shown for each:

- **Global only** (the local database made no call, so the global call is kept):
  the six checklist-demo genera from the main tutorial, `Abudefduf saxatilis` /
  `taurus`, `Prionurus laticlavius` / `biafraensis`, `Archosargus rhomboidalis`,
  and `Conodon nobilis`. With nothing to reconcile, they pass through unchanged.
- **Both agree, so the call is kept**: `Lutjanus kasmira` and `Caranx
  melampygus`. Both databases name the same species, and with `Local_advantage =
  TRUE` (the default) the tie goes to local. The species is kept.
- **Disagree and the global percent identity wins, so REGATTA downgrades to the
  LCA** (`global_lca_to_local`): `Epinephelus bontoides` (global) vs. `E. merra`
  (local), and `Mugil curema` (global) vs. `M. thoburni` (local). The global call
  has the higher identity, but because the two conflict at species, REGATTA
  reports the lowest common ancestor, genus `Epinephelus` and genus `Mugil`.
- **Disagree and the local percent identity wins, so the local call is kept**:
  `Etrumeus teres` (local, 99.4%) over `E. micropus` (global, 98.9%), and
  `Cubiceps pauciradiatus` (local) over `C. squamiceps` (global). Here the
  locally curated database has the better match, so its call is trusted outright.

Feed the reconciled result straight into the checklist step, exactly as in the
single-classifier workflow:

```{r demo-checklist}
checklist <- readRDS(system.file("extdata", "galapagos_fish_checklist.rds",
                                 package = "REGATTA"))
post <- reconcile_checklist(rec$result, checklist, output_dir = NULL)
```

## Same 12 ASVs, both methods, side by side

Whether a second database is worth it comes down to what changes versus the
single-classifier run, and the changes go both ways: some calls become less
specific (a conflict caught), and some become more specific (a regional species
recovered).

Run the main-tutorial method (obitools straight to the checklist) on the same
inputs and compare the final call:

```{r compare}
single <- reconcile_checklist(global, checklist, output_dir = NULL)

# the most specific non-NA rank per ASV, named by ASV_id (the reconcile steps
# may reorder rows, so we key on the id rather than trusting row position)
lowest <- function(r) {
  ranks <- c("domain","phylum","class","order","family","genus","species")
  v <- apply(r$result[, ranks], 1, function(x) {nz <- which(!is.na(x)); if (length(nz)) x[max(nz)] else NA})
  stats::setNames(v, r$result$ASV_id)
}
ids <- global$ASV_id
data.frame(
  obitools_call = global$species,
  single_db     = lowest(single)[ids],
  two_db        = lowest(post)[ids],
  row.names     = NULL
)
```

Nine of the twelve land in the same place, but three change, each in a different
way:

- `Mugil curema`: single-DB keeps the species (`M. curema` is on the checklist);
  two-DB drops to genus `Mugil` because the local database's `M. thoburni`
  contradicts it. Two databases caught a conflict one could not.
- `Etrumeus micropus`: single-DB keeps the global species; two-DB replaces it
  with `Etrumeus teres`, the local database's higher-identity, regionally curated
  call.
- `Cubiceps squamiceps`: single-DB downgrades to genus `Cubiceps` (`squamiceps`
  is not on the checklist); two-DB keeps the species `Cubiceps pauciradiatus`,
  because the local database's call is on the checklist. Here the second database
  makes the result more specific, not less.

## Inspecting and tuning

The reconciliation is fully auditable and adjustable. You can see why each ASV
got its call, and change how ties and percent-identity scales are handled to fit
your databases.

- The full Stage 1 audit (`reconcile_global_local/` on disk, or `rec$tracking`)
  groups its columns per field as `<field>_global` / `<field>_local` /
  `<field>_regatta` (the reconciled value). Each of the seven ranks and `pct_id`
  read global / local / regatta side by side, followed by the decision columns
  (`best_pctid_winner`, `preferred_database`, and so on) and any remaining
  classifier columns.
- `Local_advantage = FALSE` makes ties go to the global database instead.
- Percent-identity scales are auto-detected per input (obitools' 0–1
  `BEST_IDENTITY` is rescaled to 0–100 to match vsearch). Override with
  `global_pct_id_scale` / `local_pct_id_scale` if needed.

See `?reconcile_global_local` for the full argument list, and
`vignette("REGATTA-tutorial")` for the checklist-building steps this workflow
assumes.
