---
title: "REGATTA tutorial: reconciling eDNA assignments against a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What REGATTA does, and what you give it

Your classifier confidently assigns eDNA reads to species that don't live in
your study area — within the group your primer targets (e.g. a freshwater fish
called from marine MiFish data). The DNA isn't wrong; the species-level *name*
is geographically implausible. This happens when the true local species is
**absent from the reference database**, so the read matches its nearest
sequenced relative (often from the other side of the world) at the amplified
region — and it can also reflect lab contamination or degraded input DNA. Such
a call is **not** off-target amplification (it's still a fish); it is a
geographically spurious species-level assignment.

REGATTA reconciles those calls against a regional species checklist,
downgrading the taxonomic specificity of any assignment to a taxon with no
regional record — with **no hand-curated reference database** and **no blanket
percent-identity cutoff** that would
throw away good species-level calls. It walks each assignment up the taxonomic
tree only as far as the regional checklist forces, and no further. (A genuinely
*off-target* amplification — a non-target group, such as an invertebrate picked
up by a fish primer — is a separate phenomenon; running one checklist per
target group, below, keeps those visible too.)

You give REGATTA three things:

1. **The taxonomic group(s)** your primer targets, as taxon names.
2. **A regional polygon** — a WKT `POLYGON((...))` string. Draw it at
   [wktmap.com](https://wktmap.com) and copy the generated text.
3. **A classifier results file** — whatever your taxonomy-assignment step
   produced (obitools `.tab`, vsearch, Kraken2/BestTaxon, …). The format is
   auto-detected.

REGATTA is organized as three modules. This tutorial covers the first two:
**Module 1** — build the regional species checklist (run it on its own for a
species list, or as the prequel to the rest) — and **Module 2** — reconcile one
classifier's output against that checklist. **Module 3** (reconciling a global +
a local database before the checklist) is in
`vignette("REGATTA-two-database")`.

It runs on the package's worked example — **Galapagos reef fish from the MiFish
primer, classified with obitools against the global NCBI database.** Nothing in
the pipeline is fish-, marine-, or Galapagos-specific; swap in your own group,
polygon, and results file and the calls are identical.

---

## The taxonomy database — set this up first

Before anything else, one piece of infrastructure underlies the whole workflow:
a local NCBI taxonomy snapshot, built by `taxonomizr`. **Building the regional
checklist** (next) and **resolving real classifier output** (taxIDs or
scientific names) to lineages both run against it, so it is the true first step
— not an afterthought. By default REGATTA keeps one in a **persistent per-user
cache** (`tools::R_user_dir("REGATTA", "cache")`), shared across your projects
and sessions, building only the lightweight names+nodes (a few hundred MB, a few
minutes) on first use — not the multi-GB accession data. Point `sql_path` at an
existing database to reuse one you already have.

The database is a **snapshot**: REGATTA never rebuilds it silently (that would
make results irreproducible), reports its build date, and records it in
`cl$methods`, so the taxonomy version travels with your methods. Pass
`overwrite_taxonomy_files = TRUE` to refresh it in place, or `sql_path = NULL` to
skip taxonomizing at checklist-build time and defer it to `run_regatta()`.

```{r show-cache, eval = FALSE}
# Where the cache lives (built once on first use, reused across projects):
tools::R_user_dir("REGATTA", "cache")
```

(The runnable demo at the end of this tutorial needs no database — it uses a
pre-taxonomized checklist bundled with the package.)

---

# Module 1 — Build the regional species checklist

This module assembles a taxonomized regional species list from public
biodiversity databases (and any local checklists you have). Its output feeds
Module 2 below — but it is also useful on its own if you just want a clean
species list for a region × group.

## `resolve_taxa()`

**Why this matters:** every taxon your primer can amplify needs to be on the
checklist you build, because the checklist feeds *two* downstream steps and a
missing group breaks both:

- it's the **species list you hand to a reference-database builder** (e.g.
  CRABS) to download FASTA sequences and assemble your local reference
  database — a group you leave out simply never gets references downloaded; and
- it's the **regional taxa list for the geographic–taxonomic reconciliation**
  of your taxonomy table — gaps there leave real local species unrepresented,
  so a present species can't be confirmed and may be wrongly downgraded as
  having no regional record.

**How:** enter the scientific name(s) of your group. Scientific names work
directly when OBIS and GBIF both recognize them — e.g. `Mammalia`:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("Mammalia")
#>      input alias_used aphia_id valid_name  rank  kingdom worms_status n_candidates gbif_key gbif_usable     note
#> 1 Mammalia       <NA>     1837   Mammalia Class Animalia           ok            1      359        TRUE resolved
```

Some scientific names aren't matched cleanly, though — they can error or come
back empty (the ray-finned-fish classes have no usable GBIF backbone key), and
a few are ambiguous (`Vertebrata` is also a red-algae genus). For those common
cases we ship quoted common-name shortcuts that expand to the right set and
sidestep the pitfalls:

- `"fish"` → ray-finned fishes + sharks & rays + hagfishes + lampreys (the
  usual MiFish target)
- `"vertebrates"` → the vertebrate classes (the `"fish"` set plus `Mammalia`,
  `Aves`, `Reptilia`, `Amphibia`), so it works in both OBIS and GBIF — unlike
  the bare subphylum `Vertebrata`, which GBIF can't resolve to a usable key

Run `resolve_taxa()` on your names first to see what each resolves to — and
whether GBIF has a usable key — before kicking off a long download.

---

## `build_regional_checklist()`

**Why this matters:** REGATTA judges every assignment against the species that
actually occur in your region, so the first job is to assemble that list — once
per region × group. The more complete it is, the fewer real local species get
mistaken for geographically spurious calls and downgraded downstream (and the same species list also seeds
your reference database).

**How:** one call pulls the records, merges them, taxonomizes the LCA list
(against the cached taxonomy DB, or one you point it at), and **returns**
everything — and also writes it to the **required** `output_dir`, so the
expensive artifacts persist (you don't rebuild them). OBIS runs by default (the
more complete marine source); GBIF is **off by default** because its
`occ_download` is slow and asynchronous — switch it on with `GBIF = TRUE` for a
fresh download, or hand it a finished download's key/object to reuse one.
Optionally fold in your own local checklist with `CSV =` (a path anywhere on
disk; ideally a citable list). `region` and `label` name the output files;
`taxa` is the query. This needs internet (and a GBIF account if `GBIF` is on),
so it's shown, not run.

> **Turning on GBIF (`GBIF = TRUE`).** A fresh GBIF run submits an
> `occ_download` job and waits for GBIF to assemble it server-side — this
> **typically takes several minutes** (longer for large regions or broad
> taxa), so expect the call to block while it polls. GBIF requires a free
> account, and `rgbif` authenticates with your account **credentials** (there
> is no separate "API key" for downloads). Register at
> [gbif.org](https://www.gbif.org/user/profile), then store three values in
> your `~/.Renviron` so REGATTA can use them without hardcoding secrets:
>
> ```
> GBIF_USER=your_gbif_username
> GBIF_PWD=your_gbif_password
> GBIF_EMAIL=the_email_you_registered_with
> ```
>
> Edit the file with `usethis::edit_r_environ()`, save, and **restart R** so
> the variables load. Once a download finishes you can reuse it instead of
> re-running: pass its key string or the `occ_download` object to `GBIF =`.
>
> **The GBIF citation is part of your pipeline's reproducibility.** Whenever
> GBIF is used — a fresh `GBIF = TRUE` download **or** a reused download key —
> GBIF mints a citation (with a DOI) you must include in any publication using
> the data. REGATTA folds the download **key**, **DOI**, and full **citation**
> into `cl$methods` (and the written `..._methods.txt`), so the GBIF provenance
> travels with the rest of your methods rather than as a separate file. The DOI
> is fetched from the download key via `rgbif::occ_download_meta()`, so it is
> recorded even when you reuse an existing download. The key in `cl$methods`
> also lets you (or a reviewer) reproduce the exact download later via
> `GBIF = "<download_key>"`.

```{r build, eval = FALSE}
poly <- "POLYGON ((-117.421875 31.952162, -91.933594 -6.315299, -81.386719 -6.315299, -76.113281 7.710992, -82.089844 8.581021, -87.011719 13.581921, -104.238281 20.303418, -112.5 32.249974, -117.421875 31.952162))"  # E. Tropical Pacific incl. Galapagos

cl <- build_regional_checklist(
  region        = "galapagos",   # names the output files
  label         = "fish",        # ...short, even if `taxa` is a long name list
  taxa          = "fish",        # query: "fish" = ray-finned + sharks/rays/hagfish/lampreys
  regional_poly = poly,
  OBIS          = TRUE,          # primary source (default)
  GBIF          = FALSE,         # off by default; TRUE = fresh download, or a key/object to reuse one
  CSV           = "2016Aug24_Tirado-Sanchez_et_al_Galapagos_Pisces_Checklist.csv",
                                               # optional: your own checklist, ideally a
                                               # citable published list (here, a real one)
  marine        = TRUE, terrestrial = FALSE,   # keep marine + diadromous fish, drop land contaminants
  output_dir    = "local_database_checklist",             # REQUIRED: where the outputs are written
  sql_path      = file.path(tools::R_user_dir("REGATTA", "cache"), "accessionTaxa.sql")
  # ^ this IS the default (a persistent cache, built on first use; run that
  #   file.path(...) to see where). Replace with "/path/to/your/accessionTaxa.sql"
  #   to reuse an existing DB; overwrite_taxonomy_files = TRUE refreshes a stale
  #   one; sql_path = NULL skips taxonomizing (defer to run_regatta()).
)
```

> **The taxonomy database (`sql_path`).** This is the snapshot from the first
> section above. The `sql_path` shown is the default cache; point it at an
> existing DB to reuse one, pass `overwrite_taxonomy_files = TRUE` to refresh it,
> or `sql_path = NULL` to skip taxonomizing here and defer it to the reconcile
> step. When a needed DB is missing, REGATTA prompts to build it (interactive)
> or, in a script, errors with the one-line build command.

`build_regional_checklist()` **returns** a list (and writes the same outputs to
a dated `<region>_<label>_<date>` subfolder of `output_dir`, so successive runs
don't pile up — the GBIF `.zip` is kept there too):

* `cl$for_making_localdb` — a bare character vector of the unique species
  binomials (no columns), **also written one-per-line** to
  `<region>_<label>_list_for_making_localdb.txt` in the output folder. That file
  is the ready-to-use input for FASTA-building tools like
  [CRABS](https://github.com/gjeunen/reference_database_creator), which expect
  one scientific name per line — no need to write it out yourself.

* `cl$checklist_summary` — the full taxonomized table, including how each name
  resolved (scientific name / synonym / unresolved). Use it to audit the
  checklist; `NULL` if no `taxonomizr` database was available.
* `cl$for_LCA` — the `taxID` column plus the 7 taxonomic ranks, cleaned and
  ready to hand straight to `run_regatta(checklist = cl$for_LCA)`.
* `cl$methods` — a ready-to-adapt **methods/provenance sentence**: the region
  and label (and the queried taxa if they differ from the label), which sources
  were used, the date, the WKT polygon, the **NCBI taxonomy snapshot date**, and
  the citations to include (REGATTA, the OBIS/GBIF R clients, the OBIS/GBIF
  databases). **Any GBIF download key, DOI, and citation live here too** — there
  is no separate GBIF element. The R-package citations are pulled live from
  `citation()` and the GBIF line uses your download's own DOI — nothing is
  invented, but **verify each reference against your own reference manager**
  before submission.

---

## `taxonomize_checklist()` (runs in the background)

**Why this matters:** the reconciliation walks each assignment up the ranks, so
every checklist species needs its full lineage; the lookup also updates older
names to current canonical NCBI taxonomy, so a synonym on either side still
matches. You don't normally call this — `build_regional_checklist()` runs it for
you, and if you hand `run_regatta()` / `reconcile_checklist()` a checklist that
*wasn't* taxonomized, they **warn and taxonomize it on the fly**.

**How (advanced / direct use only):** call it yourself just to inspect or cache
the lineage table. The first run builds the local `taxonomizr` names+nodes
database if it isn't already there (~a few hundred MB, a few minutes — see the
first section); later runs reuse it.

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "local_database_checklist/galapagos_fish_list_for_LCA.txt",
  sql_path = "/path/to/accessionTaxa.sql"
)
```

---

# Module 2 — Reconcile your classifier against the checklist

With the Module 1 checklist in hand, this module takes your taxonomy-assignment
software's output and walks each call up the tree only as far as the regional
checklist forces — so the eDNA assignment and the regional range limits agree.

## `run_regatta()`

**Why this matters:** once the checklist is taxonomized, one call turns your raw
classifier file into a reconciled, audited taxonomy table — you don't wire the
preprocessing, reconciliation, and reporting steps together yourself.

**How:** point `run_regatta()` at your results file and the checklist from the
previous step (`cl$for_LCA`). It detects the format (here an obitools `.tab`,
recognized by its `TAXID` column), resolves the taxIDs to lineages, runs the
checklist LCA, and **returns** the reconciled table + summary. (If `cl$for_LCA`
wasn't taxonomized — no DB at build time — `run_regatta()` warns and
taxonomizes it on the fly.)

```{r run, eval = FALSE}
res <- run_regatta(
  input     = "MiFish_obitools.tab",   # obitools .tab (global DB)
  checklist = cl$for_LCA,              # from build_regional_checklist() above
  out_dir   = "regatta_out",           # REQUIRED: outputs go in a dated subfolder
  region    = "galapagos", label = "fish"  # name the run subfolder (match build)
  # sql_path defaults to the cache; pass a path to override.
)
#> returns res$taxonomy_table (the final reconciled table), res$tracking (the
#> per-ASV audit), and res$summary (per-stage stats); writes reconcile_checklist/*,
#> regatta_summary.csv, run_log.txt into regatta_out/galapagos_fish_<date>/
```

On this dataset, of the 542 obitools-assigned ASVs (out of 1,438 total):

| match rank | ASVs |
|---|---|
| species | 287 |
| genus   | 229 |
| family  | 16 |
| order   | 1 |
| class   | 9 |

The 229 genus-level and 9 class-level results are exactly the geographically
spurious calls REGATTA exists to reconcile: globally-confident species
assignments to fish with no Galapagos record, each downgraded to the lowest
rank the regional checklist can actually support.

> **Classified the same ASVs against a second, locally-curated database too?**
> REGATTA can reconcile the two classifier runs against each other before the
> checklist step — see `vignette("REGATTA-two-database")`.

---

## `reconcile_checklist()`

**Why this matters:** this is the core step `run_regatta()` calls for you — and
because it's pure base R, you can watch it work right here, with no internet and
no database, on real data bundled with the package. Seeing each ASV's outcome
makes the reconciliation concrete before you run it on your own table.

**How:** pass a taxonomy table and the taxonomized checklist. The demo below
uses the bundled Galapagos checklist and **12 real obitools assignments (10
genera)** chosen to show every outcome; the two-database vignette runs these
*same 12 ASVs* through the optional step so you can compare the methods. You
would normally pass your *whole* classifier table, not a hand-picked slice.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file(
  "extdata", "galapagos_fish_checklist.rds", package = "REGATTA"))

# 12 real obitools-assigned ASVs from the worked dataset (ASV_id + 7 ranks + pct_id).
tax <- read.csv(system.file(
  "extdata", "mifish_obitools_example.csv", package = "REGATTA"),
  stringsAsFactors = FALSE)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)
```

`$tracking` records *why* each row changed — the input species, the percent
identity, and the rank at which it matched the regional checklist:

```{r demo-tracking}
res$tracking[, c("before_species", "pct_id", "regatta_match_rank",
                 "after_genus", "after_family")]
```

The checklist step has **three outcomes**, and all three appear here:

* **Kept at species** (the call is on the checklist) — *Abudefduf saxatilis*,
  *Prionurus laticlavius*, *Lutjanus kasmira*, *Caranx melampygus*,
  *Mugil curema*, *Etrumeus micropus*.
* **Downgraded to genus** (species off the checklist, genus on) —
  *Abudefduf taurus* → *Abudefduf*, *Prionurus biafraensis* → *Prionurus*,
  *Epinephelus bontoides* → *Epinephelus*, *Cubiceps squamiceps* → *Cubiceps*.
* **Downgraded to family** (species *and* genus off the checklist) —
  *Archosargus rhomboidalis* → **Sparidae**, *Conodon nobilis* → **Haemulidae**.

The two *within-genus splits* show why this beats a flat percent-identity
cutoff. *Abudefduf saxatilis* and *A. taurus* **both matched the global
database at 100% identity**, yet only *saxatilis* is on the Galapagos
checklist, so *taurus* drops to genus — no cutoff could ever separate them.
Likewise *Prionurus laticlavius* (the Galapagos sawtail surgeonfish) stays at
species while *P. biafraensis* — a **West African** congener confidently called
at 98.8% — drops to genus. The global database simply has no reference for some
Galapagos species, so it names the nearest relative it has; REGATTA keeps the
reliable genus and drops the unreliable species.

`$stats` summarizes the before→after transition for this slice — how many calls
were kept, downgraded, or dropped, with the downgrades broken down by rank pair:

```{r demo-stats}
res$stats
```

> Six of these ASVs were *kept at species* here — but three of those calls
> change once a second, locally-curated classifier weighs in.
> `vignette("REGATTA-two-database")` runs the same 12 ASVs through that step and
> compares the two methods side by side.

---

## Interpreting the outputs

**Why this matters:** every REGATTA step that produces a taxonomy table hands
back the same three things, so results drop straight into phyloseq or MetabaR
*and* stay fully traceable — you can always show why a call changed.

**How:** read the three elements (and, when writing to disk, the matching three
CSVs per stage):

| Element | Contents |
|---|---|
| `$result`   | The **exchange format**: the id column, then `pct_id` (when the input carries one, normalized to a 0–100 percent scale for downstream filtering), then the 7 lowercase ranks. For a phyloseq `tax_table()` or MetabaR MOTU table, drop the `pct_id` column first. |
| `$tracking` | Per-ASV audit: before/after at each rank, `regatta_match_rank`, `regatta_reason` (`kept` / `non-local (geographic)` / `off-target (taxonomic)`, when the checklist carries a target group), and your passthrough columns. |
| `$stats`    | Step-level counts — the before→after transition (kept / downgraded / dropped) plus the per-rank-pair downgrade breakdown above. |

## Where to go next

* **A different results format?** `run_regatta()` auto-detects vsearch (LCA +
  `--userout`) and Kraken2/BestTaxon-style name tables. For a table whose
  per-ASV id column isn't `ASV_id` (e.g. a sequence `Hash`), pass
  `id_col = "Hash"`; a long per-sample table is collapsed to one row per ASV
  automatically. See `?run_regatta`.
* **Two reference databases? (Module 3)** If you classified the same ASVs
  against both a global and a locally-curated database, REGATTA can reconcile the
  two classifier runs against each other before the checklist step. See
  `vignette("REGATTA-two-database")` and `?reconcile_global_local`.
* **One pass per primer.** Run the pipeline once per taxonomic group your
  primers target. This also keeps genuine off-target amplifications (a
  non-target group, e.g. an invertebrate picked up by a fish primer) visible;
  mixing groups into one megalist lets them pass unflagged.
* **Pre-check your taxon names** with `resolve_taxa()` before a long download,
  and see the README troubleshooting section for GBIF/OBIS specifics.
