---
title: "REGATTA tutorial: reconciling eDNA assignments against a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

REGATTA helps reconcile eDNA taxonomic assignments with the species that are
actually known from your study region.

This is useful when a classifier assigns reads to a species that is
taxonomically plausible but geographically unlikely. For example, a fish primer
may produce a confident species-level fish assignment, but the assigned species
may not occur in your region. This often happens when the true local species is
missing from the reference database, so the read is assigned to the closest
sequenced relative available. It can also reflect contamination or low-quality
input DNA.

REGATTA addresses this by comparing classifier output against a regional species
checklist. If an assigned species has no regional record, REGATTA walks the
assignment up the taxonomic tree only as far as needed. Species-level calls are
retained when the species is regionally supported; otherwise, the call is
downgraded to the lowest supported rank, such as genus or family.

This approach preserves useful taxonomic information without relying on a blanket
percent-identity cutoff that could discard valid species-level calls.

## What you provide

You provide REGATTA with three inputs:

1. The taxonomic group or groups targeted by your primer, as taxon names.
2. A regional polygon, supplied as a WKT `POLYGON((...))` string. You can draw
   one at [wktmap.com](https://wktmap.com/) and copy the generated text.
3. A classifier results file from your taxonomy-assignment step, such as
   obitools, vsearch, Kraken2, or a similar output table.

If your classifier format is not currently supported, please open an issue so we
can improve compatibility.

## Tutorial overview

REGATTA is organized into three modules. This tutorial covers the first two:

- **Module 1:** Build a regional species checklist.
- **Module 2:** Reconcile one classifier output against that checklist.

Module 3, which reconciles global and local database classifications before the
checklist step, is covered separately in `vignette("REGATTA-two-database")`.

The worked example uses Galapagos reef fish amplified with the MiFish primer and
classified with obitools against the global EMBL database. The workflow is not
specific to fish, marine systems, or the Galapagos. You can use the same
functions with your own taxonomic group, polygon, and classifier output.

---

# Set up the taxonomy database first

REGATTA uses a local NCBI taxonomy snapshot built by `taxonomizr`. This taxonomy
database supports both checklist building and classifier reconciliation, so we
recommend setting it up before running the workflow.

By default, REGATTA stores one taxonomy database in a persistent per-user cache:

```{r show-cache, eval = FALSE}
tools::R_user_dir("REGATTA", "cache")
```

This cache is shared across projects and R sessions. On first use, REGATTA builds
the lightweight names-and-nodes database, which is usually a few hundred MB and
takes a few minutes. It does not build the multi-GB accession database unless you
explicitly provide or create one.

You can also point REGATTA to an existing taxonomy database:

```{r sql-existing, eval = FALSE}
sql_path <- "/path/to/your/accessionTaxa.sql"
```

The taxonomy database is treated as a reproducible snapshot. REGATTA does not
rebuild it silently. Instead, it reports the taxonomy build date and records it
in `cl$methods`, so the taxonomy version travels with your methods.

To refresh the cached database, use `overwrite_taxonomy_files = TRUE`. To skip
taxonomizing during checklist building and defer it to `run_regatta()`, use
`sql_path = NULL`.

The runnable demo at the end of this tutorial does not require a taxonomy database
because it uses a pre-taxonomized checklist bundled with the package.

---

# Module 1: Build the regional species checklist

Module 1 builds a taxonomized species list for your region and target group. This
checklist is used in Module 2, but it is also useful on its own if you need a
clean regional species list for reference database construction or biodiversity
summaries.

## `resolve_taxa()`

Use `resolve_taxa()` to check how REGATTA interprets your target taxon names
before starting a long checklist download.

This matters because the checklist should include every group your primer can
amplify. Missing groups can affect both downstream steps:

- Reference-database construction. If a group is missing from the checklist,
  tools such as CRABS will not download references for that group.
- Geographic reconciliation. If a real local species is missing from the
  checklist, REGATTA may be forced to downgrade a valid assignment because the
  species has no regional record in the checklist.

Scientific names work directly when both OBIS and GBIF recognize them. For
example:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("Mammalia")
#>      input alias_used aphia_id valid_name  rank  kingdom worms_status n_candidates gbif_key gbif_usable     note
#> 1 Mammalia       <NA>     1837   Mammalia Class Animalia           ok            1      359        TRUE resolved
```

Some names are not matched cleanly. For example, some ray-finned fish classes do
not have usable GBIF backbone keys, and some names are ambiguous. `Vertebrata`,
for example, is also a red-algae genus.

For common cases, REGATTA includes quoted shortcuts that expand to the
appropriate target groups:

- `"fish"`: ray-finned fishes, sharks and rays, hagfishes, and lampreys. This is
  the usual MiFish target set.
- `"vertebrates"`: the `"fish"` set plus Mammalia, Aves, Reptilia, and Amphibia.
  This works in both OBIS and GBIF, unlike the bare subphylum name `Vertebrata`.

Before starting a large download, run `resolve_taxa()` on your planned names to
confirm how they resolve and whether GBIF has a usable key.

---

## `build_regional_checklist()`

Use `build_regional_checklist()` to assemble the regional checklist.

REGATTA compares every classifier assignment against the species known from your
region. The more complete this checklist is, the fewer true local species will be
mistaken for geographically spurious assignments and downgraded later.

`build_regional_checklist()` pulls biodiversity records, merges sources,
taxonomizes the least-common-ancestor list, returns the resulting objects, and
writes the outputs to disk. Writing outputs is intentional: checklist building
can be expensive, and the written files let you reuse the artifacts without
rebuilding them.

OBIS runs by default and is often the more complete source for marine taxa. GBIF
is off by default because `occ_download()` is slower and asynchronous. You can
switch GBIF on with `GBIF = TRUE`, or reuse a completed GBIF download by passing
its key or download object to `GBIF =`.

You can also include your own regional checklist with `CSV =`. A local checklist
is especially useful when you have a citable regional list or expert-curated
records that may not be complete in public databases.

```{r build, eval = FALSE}
poly <- "POLYGON ((-117.421875 31.952162, -91.933594 -6.315299, -81.386719 -6.315299, -76.113281 7.710992, -82.089844 8.581021, -87.011719 13.581921, -104.238281 20.303418, -112.5 32.249974, -117.421875 31.952162))"

cl <- build_regional_checklist(
  region        = "galapagos",
  label         = "fish",
  taxa          = "fish",
  regional_poly = poly,
  OBIS          = TRUE,
  GBIF          = FALSE,
  CSV           = "2016Aug24_Tirado-Sanchez_et_al_Galapagos_Pisces_Checklist.csv",
  marine        = TRUE,
  terrestrial   = FALSE,
  output_dir    = "local_database_checklist",
  sql_path      = file.path(tools::R_user_dir("REGATTA", "cache"), "accessionTaxa.sql")
)
```

In this example:

- `region` names the output files.
- `label` gives the run a short readable label.
- `taxa` defines the taxonomic query.
- `regional_poly` defines the study region.
- `OBIS` and `GBIF` control public biodiversity data sources.
- `CSV` adds an optional local checklist.
- `marine` and `terrestrial` help filter records by habitat.
- `output_dir` is required and tells REGATTA where to write outputs.
- `sql_path` points to the taxonomy database. The path shown above is the default
  cache location.

### Using GBIF

A fresh GBIF run submits an `occ_download()` job and waits while GBIF assembles
the download server-side. This usually takes several minutes, and larger regions
or broad taxonomic groups may take longer.

GBIF requires a free account. `rgbif` authenticates using three values stored in
your `~/.Renviron` file:

```
GBIF_USER=your_gbif_username
GBIF_PWD=your_gbif_password
GBIF_EMAIL=the_email_you_registered_with
```

You can edit this file from R with:

```{r edit-renviron, eval = FALSE}
usethis::edit_r_environ()
```

Save the file and restart R so the variables load.

After a GBIF download finishes, you can reuse it instead of starting a new
download by passing its key to `GBIF =`, for example `GBIF = "<download_key>"`.

When GBIF is used, REGATTA records the download key, DOI, and full citation in
`cl$methods` and in the written methods file. This keeps GBIF provenance with the
rest of the workflow rather than in a separate file.

### Checklist outputs

`build_regional_checklist()` returns a list and writes the same outputs to a
dated subfolder of `output_dir`, using the pattern `<region>_<label>_<date>/`.

The main returned elements are:

| Element | Contents |
|---|---|
| `cl$for_making_localdb` | A character vector of unique species binomials. This is also written one name per line for use with FASTA-building tools such as CRABS. |
| `cl$checklist_summary` | The full taxonomized checklist, including how each name resolved. This is useful for auditing. It is `NULL` if no taxonomy database was available. |
| `cl$for_LCA` | The taxID column plus seven taxonomic ranks, cleaned and ready to pass to `run_regatta(checklist = cl$for_LCA)`. |
| `cl$methods` | A ready-to-adapt methods and provenance record, including region, label, queried taxa, sources, date, WKT polygon, taxonomy snapshot date, and relevant citations. Verify each reference against your own reference manager before submission. |

---

## `taxonomize_checklist()`

You usually do not need to call `taxonomize_checklist()` directly.
`build_regional_checklist()` runs it for you, and `run_regatta()` or
`reconcile_checklist()` can taxonomize an untaxonomized checklist on the fly.

Direct use is mainly helpful if you want to inspect or cache the lineage table
yourself.

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "local_database_checklist/galapagos_fish_list_for_LCA.txt",
  sql_path = "/path/to/accessionTaxa.sql"
)
```

This step adds the full lineage needed for reconciliation. It also updates older
names to current canonical NCBI taxonomy, so synonyms can still match where
possible.

---

# Module 2: Reconcile classifier output against the checklist

Module 2 takes your classifier output and compares it with the taxonomized
regional checklist from Module 1.

If a species-level assignment is supported by the regional checklist, REGATTA
keeps it. If the species is not regionally supported, REGATTA walks the
assignment up the taxonomic tree until it reaches the lowest rank supported by
the checklist.

---

## `run_regatta()`

Use `run_regatta()` for the standard workflow. It reads your classifier output,
resolves lineages, runs the checklist reconciliation, and writes an audited
output table.

```{r run, eval = FALSE}
res <- run_regatta(
  input     = "MiFish_obitools.tab",
  checklist = cl$for_LCA,
  out_dir   = "regatta_out",
  region    = "galapagos",
  label     = "fish"
)
```

In this example:

- `input` is the classifier output. Here, it is an obitools `.tab` file
  recognized by its `TAXID` column.
- `checklist` is the taxonomized checklist from Module 1.
- `out_dir` is required and tells REGATTA where to write results.
- `region` and `label` name the run subfolder.

`run_regatta()` returns `res$taxonomy_table`, `res$tracking`, and `res$summary`,
and writes outputs to a dated `regatta_out/galapagos_fish_<date>/` subfolder.

On the worked dataset, 542 ASVs were assigned by obitools out of 1,438 total
ASVs. After REGATTA reconciliation, the assignments matched the regional
checklist at these ranks:

| Match rank | ASVs |
|---|---:|
| species | 287 |
| genus | 229 |
| family | 16 |
| order | 1 |
| class | 9 |

The genus-level and class-level results are the kinds of geographically spurious
species-level calls REGATTA is designed to reconcile. The classifier made
globally confident species assignments, but those species had no Galapagos
record. REGATTA retained the taxonomic information that the regional checklist
could support and downgraded only the unsupported specificity.

If you classified the same ASVs against both a global and a locally curated
database, see `vignette("REGATTA-two-database")`. That workflow reconciles the
two classifier runs before the checklist step.

---

## `reconcile_checklist()`

`reconcile_checklist()` is the core reconciliation step called by
`run_regatta()`. You can also run it directly.

This is useful for learning how REGATTA behaves because it can run with no
internet and no taxonomy database when using the bundled example data.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file(
  "extdata", "galapagos_fish_checklist.rds",
  package = "REGATTA"
))

tax <- read.csv(system.file(
  "extdata", "mifish_obitools_example.csv",
  package = "REGATTA"
), stringsAsFactors = FALSE)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)
```

The example uses 12 real obitools-assigned ASVs from the worked dataset. The
subset was chosen to show all three checklist outcomes.

To inspect what happened to each ASV:

```{r demo-tracking}
res$tracking[, c(
  "before_species",
  "pct_id",
  "regatta_match_rank",
  "after_genus",
  "after_family"
)]
```

The checklist step can produce three outcomes:

1. **Kept at species:** the assigned species is present on the regional
   checklist.
2. **Downgraded to genus:** the assigned species is absent, but the genus is
   regionally supported.
3. **Downgraded to family:** both the assigned species and genus are absent, but
   the family is regionally supported.

In the example:

- **Kept at species:** *Abudefduf saxatilis*, *Prionurus laticlavius*,
  *Lutjanus kasmira*, *Caranx melampygus*, *Mugil curema*, and
  *Etrumeus micropus*.
- **Downgraded to genus:** *Abudefduf taurus*, *Prionurus biafraensis*,
  *Epinephelus bontoides*, and *Cubiceps squamiceps*.
- **Downgraded to family:** *Archosargus rhomboidalis* and *Conodon nobilis*.

These examples show why regional reconciliation can outperform a flat
percent-identity cutoff. *Abudefduf saxatilis* and *Abudefduf taurus* both
matched the global database at 100% identity, but only *A. saxatilis* is on the
Galapagos checklist. REGATTA therefore keeps *A. saxatilis* at species and
downgrades *A. taurus* to genus.

Similarly, *Prionurus laticlavius* is retained at species, while
*Prionurus biafraensis*, a West African congener assigned at 98.8% identity, is
downgraded to genus. The global database may be naming the closest sequenced
relative rather than the true local species. REGATTA keeps the reliable
genus-level information and removes the unsupported species-level specificity.

You can summarize the before-and-after transition with:

```{r demo-stats}
res$stats
```

Six ASVs were kept at species in this checklist-only example. Three of those
calls change when the optional second-database reconciliation step is included.
See `vignette("REGATTA-two-database")`.

---

# Interpreting REGATTA outputs

Every REGATTA step that produces a taxonomy table returns the same three core
elements. This makes the results easier to pass into downstream tools such as
`phyloseq` or MetabaR while preserving an audit trail.

| Element | Contents |
|---|---|
| `$result` | The exchange-format taxonomy table: ID column, `pct_id` when available, and seven lowercase taxonomic ranks. For a `phyloseq::tax_table()` or MetabaR MOTU table, drop `pct_id` first. |
| `$tracking` | Per-ASV audit table showing before and after taxonomy, match rank, reconciliation reason, and passthrough columns. |
| `$stats` | Step-level counts, including kept, downgraded, dropped, and rank-pair downgrade summaries. |

The tracking table is the best place to check why a call changed. Depending on
the available checklist information, `regatta_reason` can distinguish outcomes
such as:

- `kept`
- `non-local (geographic)`
- `off-target (taxonomic)`

---

# Where to go next

## Use a different classifier format

`run_regatta()` can auto-detect several common input types, including:

- obitools `.tab`
- vsearch LCA output
- vsearch `--userout`
- Kraken2-style or BestTaxon-style name tables

If your ASV ID column has a different name, pass it explicitly with
`id_col = "Hash"`. Long per-sample tables are collapsed to one row per ASV
automatically. See `?run_regatta`.

## Reconcile global and local database results

If you classified the same ASVs against both a global database and a locally
curated database, REGATTA can reconcile those two classifier runs before applying
the regional checklist. See `vignette("REGATTA-two-database")` and
`?reconcile_global_local`.

## Run one pass per primer

Run REGATTA once per taxonomic group targeted by your primer set.

This helps keep genuine off-target amplifications visible. For example, if an
invertebrate is amplified by a fish primer, running the fish checklist separately
prevents that non-target record from passing through as part of a mixed
"megalist."

## Pre-check taxon names

Before starting a long OBIS or GBIF download, check your target names with
`resolve_taxa()`. For GBIF and OBIS troubleshooting, see the README
troubleshooting section.
