---
title: "REGATTA advanced usage: pre-checks, tuning, and formats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA advanced usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette collects optional steps and details that build on the core
workflow in `vignette("REGATTA-tutorial")`: pre-checking taxon names, taxonomy-
database options, turning on GBIF, resolving a checklist yourself, inspecting the
checklist step ASV by ASV, reading the outputs, and handling other classifier
formats. None of it is needed for a first run.

## Pre-check your taxon names with `resolve_taxa()`

The checklist should include every group your primer can amplify. A missing group
breaks two things: a reference-database builder such as CRABS never downloads
references for it, and a real local species left off the checklist can force a
valid assignment to be downgraded. `resolve_taxa()` shows how REGATTA interprets
your target names — and whether GBIF has a usable backbone key — before you start
a long download.

Scientific names work directly when both OBIS and GBIF recognize them:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("Mammalia")
#>      input alias_used aphia_id valid_name  rank  kingdom worms_status n_candidates gbif_key gbif_usable     note
#> 1 Mammalia       <NA>     1837   Mammalia Class Animalia           ok            1      359        TRUE resolved
```

Some names are not matched cleanly: a few ray-finned fish classes have no usable
GBIF backbone key, and some names are ambiguous (`Vertebrata` is also a red-algae
genus). For common cases REGATTA ships shortcuts that expand to the right groups:

- `"fish"`: ray-finned fishes, sharks and rays, hagfishes, and lampreys — the
  usual MiFish target set.
- `"vertebrates"`: the `"fish"` set plus Mammalia, Aves, Reptilia, and Amphibia.
  This works in both OBIS and GBIF, unlike the bare subphylum name `Vertebrata`.

## Taxonomy-database options

The tutorial keeps the snapshot in the working directory and builds it on first
use. You can point `sql_path` anywhere, including an `accessionTaxa.sql` you
already have, to reuse it instead of building a new one:

```{r sql-existing, eval = FALSE}
sql_path <- "/path/to/an/existing/accessionTaxa.sql"
```

The database is a reproducible snapshot: REGATTA never rebuilds it silently. It
reports the build date and records it in `cl$methods`, so the taxonomy version
travels with your methods. To refresh a stale snapshot in place, pass
`overwrite_taxonomy_files = TRUE`. To skip taxonomizing during checklist building
and defer it to `run_regatta()`, pass `sql_path = NULL`.

## Turning on GBIF

OBIS runs by default and is often the more complete marine source. GBIF is off by
default because its download is slower and asynchronous. Switch it on with
`GBIF = TRUE`, or reuse a finished download by passing its key or download object
to `GBIF =`.

A fresh GBIF run submits an `occ_download()` job and waits while GBIF assembles
it server-side — usually several minutes, longer for large regions or broad taxa.
GBIF requires a free account. `rgbif` authenticates using three values stored in
your `~/.Renviron` file:

```
GBIF_USER=your_gbif_username
GBIF_PWD=your_gbif_password
GBIF_EMAIL=the_email_you_registered_with
```

Edit that file from R, then save and restart R so the variables load:

```{r edit-renviron, eval = FALSE}
usethis::edit_r_environ()
```

Whenever GBIF is used, REGATTA records the download key, DOI, and full citation in
`cl$methods` and in the written methods file, keeping GBIF provenance with the
rest of the workflow. Verify each reference against your own reference manager
before submission.

## Resolving a checklist yourself with `taxonomize_checklist()`

You usually do not call this directly — `build_regional_checklist()` runs it for
you, and `run_regatta()` will taxonomize an untaxonomized checklist on the fly.
Call it directly only to inspect or cache the lineage table:

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "regatta_checklist/galapagos_fish_list_for_LCA.txt",
  sql_path = "accessionTaxa.sql"
)
```

It adds the full lineage each name needs for reconciliation and updates older
names to current canonical NCBI taxonomy, so synonyms still match.

## Inside the checklist step: `reconcile_checklist()`

`reconcile_checklist()` is the core step `run_regatta()` calls for you. Running it
directly shows what happens to each ASV — and, because it uses data bundled with
the package, it runs here with no internet and no taxonomy database. The demo uses
12 real obitools-assigned ASVs chosen to show all three outcomes.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file("extdata", "galapagos_fish_checklist.rds",
                                 package = "REGATTA"))
tax <- read.csv(system.file("extdata", "mifish_obitools_example.csv",
                            package = "REGATTA"), stringsAsFactors = FALSE)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)

res$tracking[, c("before_species", "pct_id", "regatta_match_rank",
                 "after_genus", "after_family")]
```

The checklist step has three outcomes, all present here:

- **Kept at species** (the call is on the checklist): *Abudefduf saxatilis*,
  *Prionurus laticlavius*, *Lutjanus kasmira*, *Caranx melampygus*,
  *Mugil curema*, *Etrumeus micropus*.
- **Downgraded to genus** (species off the checklist, genus on): *Abudefduf
  taurus*, *Prionurus biafraensis*, *Epinephelus bontoides*, *Cubiceps
  squamiceps*.
- **Downgraded to family** (species and genus off): *Archosargus rhomboidalis*,
  *Conodon nobilis*.

The two within-genus splits show why this beats a flat percent-identity cutoff.
*Abudefduf saxatilis* and *A. taurus* both matched the global database at 100%
identity, yet only *saxatilis* is on the Galapagos checklist, so *taurus* drops to
genus — no cutoff could separate them. Likewise *Prionurus laticlavius* stays at
species while *P. biafraensis*, a West African congener called at 98.8%, drops to
genus: the global database simply has no reference for some Galapagos species, so
it names the nearest relative it has.

`$stats` summarizes the before/after transition:

```{r demo-stats}
res$stats
```

For a walkthrough of what each output element contains and how to read the
tracking table, see the "Interpreting the outputs" section of
`vignette("REGATTA-tutorial")`.

## Other classifier formats

`run_regatta()` auto-detects several input types:

- obitools `.tab` (recognized by its `TAXID` column)
- vsearch LCA output
- vsearch `--userout`
- Kraken2-style or BestTaxon-style name tables

If your ASV id column has a different name, pass it explicitly with
`id_col = "Hash"`. Long per-sample tables are collapsed to one row per ASV
automatically. See `?run_regatta`.

## One pass per primer

Run REGATTA once per taxonomic group your primers target. This keeps genuine
off-target amplifications visible: if an invertebrate is amplified by a fish
primer, running the fish checklist separately prevents that non-target record
from passing through as part of a mixed "megalist."
