---
title: "Adding a second classifier: global vs. local database reconciliation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-database workflow (global vs. local)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette is the optional add-on to `vignette("REGATTA-tutorial")`. Start
there for building and taxonomizing the regional checklist and the basic
single-classifier workflow; here we cover the extra step for when you have
**the same ASVs classified against two reference databases**.

## When you'd use this

A common eDNA setup classifies each ASV twice:

- against a **global** reference database (e.g. obitools + NCBI/EMBL), which has
  broad coverage but confidently names species from anywhere in the world, and
- against a **locally-curated** database (e.g. vsearch+SINTAX + a regional
  reference), which is sparser but regionally trustworthy.

`reconcile_global_local()` reconciles the two **before** the checklist step:
per ASV it keeps the call with the higher percent identity, and when the global
call wins but the local database disagrees, it downgrades to the lowest common
ancestor of the two — so a species-level conflict becomes an honest
genus-level call instead of silently trusting the global database.

It's the **same Galapagos MiFish dataset** as the main tutorial: the identical
1,438 ASVs, now classified a second way.

## The one-call path

Pass both results to `run_regatta()` as a named `list(global =, local =)`. It
runs `reconcile_global_local()`, then `reconcile_checklist()`, in one call;
formats are auto-detected (here the local side is a vsearch `lca` taxonomy file
plus a `--userout` file carrying percent identities).

```{r run, eval = FALSE}
library(REGATTA)

run_regatta(
  input = list(
    global = "MiFish_obitools.tab",                 # obitools (global NCBI DB)
    local  = c("MiFish_vsearch_lca.txt",            # vsearch LCA taxonomy
               "MiFish_vsearch_userout.txt")        # vsearch %ID
  ),
  checklist = "custom_db/galapagos_fish_checklist.rds",
  sql_path  = "/path/to/accessionTaxa.sql"
)
#> -> regatta_out/{ reconcile_global_local/*, reconcile_checklist/*,
#>                  regatta_summary.csv, run_log.txt }
```

On the full dataset, of the 1,438 ASVs, 892 were unassigned by both
classifiers and 546 were assigned by at least one. Stage 1 resolved those:

| | ASVs |
|---|---|
| best %ID: global won | 403 |
| best %ID: local won | 143 |
| global downgraded to global∩local LCA | 45 |

and Stage 2 (checklist) then gave 295 species / 231 genus / 19 family / 1 order
/ 892 off-target — compare the main tutorial's single-classifier obitools run
(287 species), the difference being the local database's contribution.

## A runnable demonstration (no internet, no database)

`reconcile_global_local()` is pure base R. It runs live here on **real slices
of the Galapagos MiFish dataset** bundled with the package — 12 real ASVs
classified both ways, chosen to show each decision (you would normally pass the
whole tables).

```{r demo}
library(REGATTA)

global <- read.csv(system.file("extdata", "mifish_obitools_example.csv",
                               package = "REGATTA"), stringsAsFactors = FALSE)
local  <- read.csv(system.file("extdata", "mifish_vsearch_example.csv",
                               package = "REGATTA"), stringsAsFactors = FALSE)

rec <- reconcile_global_local(global, local, output_dir = NULL)
```

`$tracking` shows the per-ASV decision — both classifiers' calls side by side,
which won on percent identity, and the database REGATTA preferred:

```{r demo-tracking}
rec$tracking[, c("ASV_id", "species_global", "pct_id_global",
                 "species_local", "pct_id_local",
                 "best_pctid_winner", "preferred_database",
                 "preferred_scientific_name")]
```

Reading the rows, three outcomes appear:

* **Global only** — *Abudefduf saxatilis* / *taurus*, *Lutjanus
  argentimaculatus*, *Caranx crysos*, *Prionurus laticlavius* / *biafraensis*,
  *Archosargus rhomboidalis*: the local database made no call, so the global
  assignment is kept (`preferred_database = global`).
* **Both agree** — *Mugil thoburni*, *Lutjanus kasmira*, *Caranx melampygus*:
  the two databases name the same species; with `Local_advantage = TRUE` (the
  default) the tie goes to local. The species is kept.
* **They disagree** — *Epinephelus bontoides* (global) vs. *Epinephelus merra*
  (local), and *Mugil curema* (global) vs. *Mugil thoburni* (local): the global
  call has the higher %ID, but because the two databases conflict at species,
  REGATTA downgrades to the lowest common ancestor — genus ***Epinephelus*** and
  genus ***Mugil*** (`preferred_database = global_lca_to_local`).

That *Mugil curema* row (`MiFish_0000222`) is the payoff. In the main tutorial's
single-classifier run, obitools' *Mugil curema* **stayed at species** because
*M. curema* is on the Galapagos checklist. Here the local database's competing
*Mugil thoburni* reveals the species-level call is not trustworthy, and the
two-database step conservatively reports genus *Mugil* instead — a downgrade the
single classifier had no way to make.

Feed the reconciled result straight into the checklist step, exactly as in the
single-classifier workflow:

```{r demo-checklist}
checklist <- readRDS(system.file("extdata", "galapagos_fish_checklist.rds",
                                 package = "REGATTA"))
post <- reconcile_checklist(rec$result, checklist, output_dir = NULL)
post$tracking[, c("ASV_id", "before_species", "regatta_match_rank", "after_genus")]
```

## Inspecting and tuning

- The full Stage 1 audit (`reconcile_global_local/` on disk, or `rec$tracking`)
  carries every input column from both classifiers suffixed `_global` /
  `_local`, plus `preferred_pctid` and the reconciled `preferred_<rank>`
  lineage — so you can see exactly why each ASV got its call.
- `Local_advantage = FALSE` makes ties go to the global database instead.
- Percent-identity scales are auto-detected per input (obitools' 0–1
  `BEST_IDENTITY` is rescaled to 0–100 to match vsearch); override with
  `global_pct_id_scale` / `local_pct_id_scale` if needed.

See `?reconcile_global_local` for the full argument list, and
`vignette("REGATTA-tutorial")` for the checklist-building steps this workflow
assumes.
