---
title: "REGATTA tutorial: reconciling eDNA assignments against a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What REGATTA does, and what you give it

Your classifier confidently assigns eDNA reads to species that don't live in
your study area, because the closest match in a global reference database is
often a relative from the other side of the world. The DNA isn't wrong — the
reference database is geographically and genetically incomplete and carries
outdated names, which together produce geographically spurious species-level
calls. REGATTA reconciles those calls against a regional species checklist,
downgrading the taxonomic specificity of any assignment to a taxon with no
regional record — with **no hand-curated reference database** and **no blanket
percent-identity cutoff** that would
throw away good species-level calls. It walks each assignment up the taxonomic
tree only as far as the regional checklist forces, and no further.

You give REGATTA three things:

1. **The taxonomic group(s)** your primer targets, as taxon names.
2. **A regional polygon** — a WKT `POLYGON((...))` string. Draw it at
   [wktmap.com](https://wktmap.com) and copy the generated text.
3. **A classifier results file** — whatever your taxonomy-assignment step
   produced (obitools `.tab`, vsearch, Kraken2/BestTaxon, …). The format is
   auto-detected.

This tutorial runs the whole pipeline on the package's worked example —
**Galapagos reef fish from the MiFish primer, classified with obitools against
the global NCBI database.** Nothing in the pipeline is fish-, marine-, or
Galapagos-specific; swap in your own group, polygon, and results file and the
calls are identical.

---

## `resolve_taxa()`

**Why this matters:** every taxon your primer can amplify needs to be on the
checklist you build, because the checklist feeds *two* downstream steps and a
missing group breaks both:

- it's the **species list you hand to a reference-database builder** (e.g.
  CRABS) to download FASTA sequences and assemble your local reference
  database — a group you leave out simply never gets references downloaded; and
- it's the **regional taxa list for the geographic–taxonomic reconciliation**
  of your taxonomy table — gaps there leave real local species unrepresented,
  so a present species can't be confirmed and may be wrongly downgraded as
  having no regional record.

**How:** enter the scientific name(s) of your group. Scientific names work
directly when OBIS and GBIF both recognize them — e.g. `Mammalia`:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("Mammalia")
#>      input alias_used aphia_id valid_name  rank  kingdom worms_status n_candidates gbif_key gbif_usable     note
#> 1 Mammalia       <NA>     1837   Mammalia Class Animalia           ok            1      359        TRUE resolved
```

Some scientific names aren't matched cleanly, though — they can error or come
back empty (the ray-finned-fish classes have no usable GBIF backbone key), and
a few are ambiguous (`Vertebrata` is also a red-algae genus). For those common
cases we ship quoted common-name shortcuts that expand to the right set and
sidestep the pitfalls:

- `"fish"` → ray-finned fishes + sharks & rays + hagfishes + lampreys (the
  usual MiFish target)
- `"vertebrates"` → the vertebrate classes (the `"fish"` set plus `Mammalia`,
  `Aves`, `Reptilia`, `Amphibia`), so it works in both OBIS and GBIF — unlike
  the bare subphylum `Vertebrata`, which GBIF can't resolve to a usable key

Run `resolve_taxa()` on your names first to see what each resolves to — and
whether GBIF has a usable key — before kicking off a long download.

---

## `build_regional_checklist()`

**Why this matters:** REGATTA judges every assignment against the species that
actually occur in your region, so the first job is to assemble that list — once
per region × group. The more complete it is, the fewer real local species get
mistaken for geographically spurious calls and downgraded downstream (and the same species list also seeds
your reference database).

**How:** one call pulls the records, merges them, taxonomizes the LCA list, and
**returns** everything — assign the result. OBIS runs by default (the more
complete marine source); GBIF is **off by default** because its `occ_download`
is slow and asynchronous — switch it on with `GBIF = TRUE` for a fresh download,
or hand it a finished download's key/object to reuse one. Fold in any local
checklist with `CSV =` (a path anywhere on disk). `region` and `label` name the
output files *if* you write them; `taxa` is the query. This needs internet (and
a GBIF account if `GBIF` is on, and a `taxonomizr` database for the background
taxonomize), so it's shown, not run.

> **Turning on GBIF (`GBIF = TRUE`).** A fresh GBIF run submits an
> `occ_download` job and waits for GBIF to assemble it server-side — this
> **typically takes several minutes** (longer for large regions or broad
> taxa), so expect the call to block while it polls. GBIF requires a free
> account, and `rgbif` authenticates with your account **credentials** (there
> is no separate "API key" for downloads). Register at
> [gbif.org](https://www.gbif.org/user/profile), then store three values in
> your `~/.Renviron` so REGATTA can use them without hardcoding secrets:
>
> ```
> GBIF_USER=your_gbif_username
> GBIF_PWD=your_gbif_password
> GBIF_EMAIL=the_email_you_registered_with
> ```
>
> Edit the file with `usethis::edit_r_environ()`, save, and **restart R** so
> the variables load. Once a download finishes you can reuse it instead of
> re-running: pass its key string or the `occ_download` object to `GBIF =`.
>
> **Keep the GBIF citation — it's part of your pipeline's reproducibility.**
> When (and only when) you run a fresh `GBIF = TRUE` download, GBIF mints a
> citation (with a DOI) that it asks you to include in any publication using the
> data. REGATTA returns it on `cl$GBIF_download_citation` (a one-row table with
> the `download_key`, `doi`, and full `citation` string) and prints it to the
> console; it does **not** silently write it to disk. Save it yourself — e.g.
> `saveRDS(cl$GBIF_download_citation, "gbif_provenance.rds")` or paste the
> `citation` into your methods — or pass `output_dir` to
> `build_regional_checklist()` to have it written to `..._GBIF_download_info.txt`
> automatically. The same `download_key` lets you (or a reviewer) reproduce the
> exact download later via `GBIF = "0042681-260519110011954"`.

```{r gbif-citation, eval = FALSE}
# After a fresh GBIF = TRUE run, preserve the key + citation:
cl$GBIF_download_citation$download_key   # reuse: build_regional_checklist(..., GBIF = <key>)
cl$GBIF_download_citation$citation       # paste this into your paper's methods
saveRDS(cl$GBIF_download_citation, "gbif_provenance.rds")
```

```{r build, eval = FALSE}
poly <- "POLYGON ((-117.421875 31.952162, -91.933594 -6.315299, -81.386719 -6.315299, -76.113281 7.710992, -82.089844 8.581021, -87.011719 13.581921, -104.238281 20.303418, -112.5 32.249974, -117.421875 31.952162))"  # E. Tropical Pacific incl. Galapagos

cl <- build_regional_checklist(
  region        = "galapagos",   # names the output files (if written)
  label         = "fish",        # ...short, even if `taxa` is a long name list
  taxa          = "fish",        # query: "fish" = ray-finned + sharks/rays/hagfish/lampreys
  regional_poly = poly,
  OBIS          = TRUE,          # primary source (default)
  GBIF          = FALSE,         # off by default; TRUE = fresh download, or a key/object to reuse one
  CSV           = "2016Aug24_Tirado-Sanchez_et_al_Galapagos_Pisces_Checklist.csv",
  marine        = TRUE, terrestrial = FALSE,   # keep marine + diadromous fish, drop land contaminants
  sql_path      = "/path/to/accessionTaxa.sql"
  # output_dir  = "my_checklists"              # uncomment to also write the files
)
```

`build_regional_checklist()` **returns** a list, writing nothing unless you
pass `output_dir`:

* `cl$for_making_localdb` — a bare character vector of the unique species
  binomials (no columns). This is the input for FASTA-building tools like
  [CRABS](https://github.com/gjeunen/reference_database_creator), which expect
  one scientific name per line. Write it out with:

    ```{r write-localdb, eval = FALSE}
    writeLines(cl$for_making_localdb, "galapagos_fish_for_crabs.txt")
    ```

* `cl$checklist_summary` — the full taxonomized table, including how each name
  resolved (scientific name / synonym / unresolved). Use it to audit the
  checklist; `NULL` if no `taxonomizr` database was available.
* `cl$for_LCA` — the `taxID` column plus the 7 taxonomic ranks, cleaned and
  ready to hand straight to `run_regatta(checklist = cl$for_LCA)`.
* `cl$methods` — a ready-to-adapt **methods/provenance sentence**: which sources
  were used, the date, the WKT polygon, and the citations to include (REGATTA,
  the OBIS/GBIF R clients, the OBIS/GBIF databases). The R-package citations are
  pulled live from `citation()` and the GBIF line uses your download's own DOI —
  nothing is invented, but **verify each reference against your own reference
  manager** before submission.
* `cl$GBIF_download_citation` — present **only when `GBIF = TRUE`** (a fresh
  download): the **download key** (reuse with `GBIF = cl$GBIF_download_citation$download_key`)
  and the **GBIF citation/DOI** you must cite. Keep this — see above.

---

## `taxonomize_checklist()` (runs in the background)

**Why this matters:** the reconciliation walks each assignment up the ranks, so
every checklist species needs its full lineage; the lookup also updates older
names to current canonical NCBI taxonomy, so a synonym on either side still
matches. You don't normally call this — `build_regional_checklist()` runs it for
you, and if you hand `run_regatta()` / `reconcile_checklist()` a checklist that
*wasn't* taxonomized, they **warn and taxonomize it on the fly**.

**How (advanced / direct use only):** call it yourself just to inspect or cache
the lineage table. The first run builds the local `taxonomizr` database (~15
min, several GB); later runs reuse it.

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "local_database_checklist/comprehensive_galapagos_fish_list_for_LCA.txt",
  sql_path = "/path/to/accessionTaxa.sql"
)
```

---

## `run_regatta()`

**Why this matters:** once the checklist is taxonomized, one call turns your raw
classifier file into a reconciled, audited taxonomy table — you don't wire the
preprocessing, reconciliation, and reporting steps together yourself.

**How:** point `run_regatta()` at your results file and the checklist from the
previous step (`cl$for_LCA`). It detects the format (here an obitools `.tab`,
recognized by its `TAXID` column), resolves the taxIDs to lineages, runs the
checklist LCA, and **returns** the reconciled table + summary. (If `cl$for_LCA`
wasn't taxonomized — no DB at build time — `run_regatta()` warns and
taxonomizes it on the fly.)

```{r run, eval = FALSE}
res <- run_regatta(
  input     = "MiFish_obitools.tab",   # obitools .tab (global DB)
  checklist = cl$for_LCA,              # from build_regional_checklist() above
  sql_path  = "/path/to/accessionTaxa.sql"
  # out_dir = "regatta_out"            # uncomment to also write the output bundle
)
#> returns res$post_checklist (reconciled table) + res$summary (21-row).
#> with out_dir set, also writes reconcile_checklist/*, regatta_summary.csv, run_log.txt
```

On this dataset, of the 542 obitools-assigned ASVs (out of 1,438 total):

| match rank | ASVs |
|---|---|
| species | 287 |
| genus   | 229 |
| family  | 16 |
| order   | 1 |
| class   | 9 |

The 229 genus-level and 9 class-level results are exactly the geographically
spurious calls REGATTA exists to reconcile: globally-confident species
assignments to fish with no Galapagos record, each downgraded to the lowest
rank the regional checklist can actually support.

> **Classified the same ASVs against a second, locally-curated database too?**
> REGATTA can reconcile the two classifier runs against each other before the
> checklist step — see `vignette("REGATTA-two-database")`.

---

## `reconcile_checklist()`

**Why this matters:** this is the core step `run_regatta()` calls for you — and
because it's pure base R, you can watch it work right here, with no internet and
no database, on real data bundled with the package. Seeing each ASV's outcome
makes the reconciliation concrete before you run it on your own table.

**How:** pass a taxonomy table and the taxonomized checklist. The demo below
uses the bundled Galapagos checklist and **12 real obitools assignments (10
genera)** chosen to show every outcome; the two-database vignette runs these
*same 12 ASVs* through the optional step so you can compare the methods. You
would normally pass your *whole* classifier table, not a hand-picked slice.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file(
  "extdata", "galapagos_fish_checklist.rds", package = "REGATTA"))

# 12 real obitools-assigned ASVs from the worked dataset (ASV_id + 7 ranks + pct_id).
tax <- read.csv(system.file(
  "extdata", "mifish_obitools_example.csv", package = "REGATTA"),
  stringsAsFactors = FALSE)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)
```

`$tracking` records *why* each row changed — the input species, the percent
identity, and the rank at which it matched the regional checklist:

```{r demo-tracking}
res$tracking[, c("before_species", "pct_id", "regatta_match_rank",
                 "after_genus", "after_family")]
```

The checklist step has **three outcomes**, and all three appear here:

* **Kept at species** (the call is on the checklist) — *Abudefduf saxatilis*,
  *Prionurus laticlavius*, *Lutjanus kasmira*, *Caranx melampygus*,
  *Mugil curema*, *Etrumeus micropus*.
* **Downgraded to genus** (species off the checklist, genus on) —
  *Abudefduf taurus* → *Abudefduf*, *Prionurus biafraensis* → *Prionurus*,
  *Epinephelus bontoides* → *Epinephelus*, *Cubiceps squamiceps* → *Cubiceps*.
* **Downgraded to family** (species *and* genus off the checklist) —
  *Archosargus rhomboidalis* → **Sparidae**, *Conodon nobilis* → **Haemulidae**.

The two *within-genus splits* show why this beats a flat percent-identity
cutoff. *Abudefduf saxatilis* and *A. taurus* **both matched the global
database at 100% identity**, yet only *saxatilis* is on the Galapagos
checklist, so *taurus* drops to genus — no cutoff could ever separate them.
Likewise *Prionurus laticlavius* (the Galapagos sawtail surgeonfish) stays at
species while *P. biafraensis* — a **West African** congener confidently called
at 98.8% — drops to genus. The global database simply has no reference for some
Galapagos species, so it names the nearest relative it has; REGATTA keeps the
reliable genus and drops the unreliable species.

`$stats` gives the match-rank distribution for this slice:

```{r demo-stats}
res$stats
```

> Six of these ASVs were *kept at species* here — but three of those calls
> change once a second, locally-curated classifier weighs in.
> `vignette("REGATTA-two-database")` runs the same 12 ASVs through that step and
> compares the two methods side by side.

---

## Interpreting the outputs

**Why this matters:** every REGATTA step that produces a taxonomy table hands
back the same three things, so results drop straight into phyloseq or MetabaR
*and* stay fully traceable — you can always show why a call changed.

**How:** read the three elements (and, when writing to disk, the matching three
CSVs per stage):

| Element | Contents |
|---|---|
| `$result`   | The **exchange format**: exactly the id column + 7 lowercase ranks. Drop straight into a phyloseq `tax_table()` or a MetabaR MOTU table. |
| `$tracking` | Per-ASV audit: before/after at each rank, `regatta_match_rank`, and your passthrough columns. |
| `$stats`    | Step-level counts — the match-rank distribution above. |

## Where to go next

* **A different results format?** `run_regatta()` auto-detects vsearch (LCA +
  `--userout`) and Kraken2/BestTaxon-style name tables. For a table whose
  per-ASV id column isn't `ASV_id` (e.g. a sequence `Hash`), pass
  `id_col = "Hash"`; a long per-sample table is collapsed to one row per ASV
  automatically. See `?run_regatta`.
* **Two reference databases?** If you classified the same ASVs against both a
  global and a locally-curated database, REGATTA can reconcile the two
  classifier runs against each other before the checklist step. See
  `vignette("REGATTA-two-database")` and `?reconcile_global_local`.
* **One pass per primer.** Run the pipeline once per taxonomic group your
  primers target. Mixing groups into one megalist lets a primer's incidental
  amplification of another group slip through unflagged.
* **Pre-check your taxon names** with `resolve_taxa()` before a long download,
  and see the README troubleshooting section for GBIF/OBIS specifics.
