---
title: "REGATTA tutorial: correcting off-target eDNA assignments with a regional checklist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REGATTA tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What REGATTA does, and what you give it

Your classifier confidently assigns eDNA reads to species that don't live in
your study area, because the closest match in a global reference database is
often a relative from the other side of the world. REGATTA corrects those
off-target calls against a regional species checklist — with **no hand-curated
reference database** and **no blanket percent-identity cutoff** that would
throw away good species-level calls. It walks each assignment up the taxonomic
tree only as far as the regional checklist forces, and no further.

You give REGATTA three things:

1. **The taxonomic group(s)** your primer targets, as taxon names.
2. **A regional polygon** — a WKT `POLYGON((...))` string. Draw it at
   [wktmap.com](https://wktmap.com) and copy the generated text.
3. **A classifier results file** — whatever your taxonomy-assignment step
   produced (obitools `.tab`, vsearch, Kraken2/BestTaxon, …). The format is
   auto-detected.

This tutorial runs the whole pipeline on the package's worked example —
**Galapagos reef fish from the MiFish primer, classified with obitools against
the global NCBI database.** Nothing in the pipeline is fish-, marine-, or
Galapagos-specific; swap in your own group, polygon, and results file and the
calls are identical.

---

## `resolve_taxa()`

**Why this matters:** every taxon your primer can amplify needs to be on the
checklist you build, because the checklist feeds *two* downstream steps and a
missing group breaks both:

- it's the **species list you hand to a reference-database builder** (e.g.
  CRABS) to download FASTA sequences and assemble your local reference
  database — a group you leave out simply never gets references downloaded; and
- it's the **regional taxa list for the geographic–taxonomic reconciliation**
  of your taxonomy table — gaps there leave real local species unrepresented,
  so a present species can't be confirmed and may be wrongly flagged
  off-target.

**How:** enter the scientific name(s) of your group. Scientific names work
directly when OBIS and GBIF both recognize them — e.g. `Mammalia`:

```{r resolve, eval = FALSE}
library(REGATTA)

resolve_taxa("Mammalia")
#>      input alias_used aphia_id valid_name  rank  kingdom worms_status n_candidates gbif_key gbif_usable     note
#> 1 Mammalia       <NA>     1837   Mammalia Class Animalia           ok            1      359        TRUE resolved
```

Some scientific names aren't matched cleanly, though — they can error or come
back empty (the ray-finned-fish classes have no usable GBIF backbone key), and
a few are ambiguous (`Vertebrata` is also a red-algae genus). For those common
cases we ship quoted common-name shortcuts that expand to the right set and
sidestep the pitfalls:

- `"fish"` → ray-finned fishes + sharks & rays + hagfishes + lampreys (the
  usual MiFish target)
- `"vertebrates"` → the vertebrate subphylum `Vertebrata`, routed to return
  vertebrates and **not** the red-algae genus of the same name

Run `resolve_taxa()` on your names first to see what each resolves to — and
whether GBIF has a usable key — before kicking off a long download.

---

## Build the regional checklist

**Why this matters:** REGATTA judges every assignment against the species that
actually occur in your region, so the first job is to assemble that list — once
per region × group. The more complete it is, the fewer real local species get
mistaken for off-target calls downstream (and, as above, the same list is what
seeds your reference database).

**How:** pull species inside your polygon from OBIS (the more complete marine
source) with `OBIS_download()` and from GBIF (a supplement) with
`GBIF_download()`, then merge them — plus any local checklist you already
have — with `build_regional_checklist()`. These calls need internet and a GBIF
account (see the README "Setup" section), so they're shown but not run here.

```{r build, eval = FALSE}
poly <- "POLYGON ((-117.421875 31.952162, -91.933594 -6.315299, -81.386719 -6.315299, -76.113281 7.710992, -82.089844 8.581021, -87.011719 13.581921, -104.238281 20.303418, -112.5 32.249974, -117.421875 31.952162))"  # E. Tropical Pacific incl. Galapagos

# OBIS — primary regional source. "fish" expands to ray-finned fishes +
# elasmobranchs; OBIS is queried by AphiaID, so any fish synonym works.
OBIS_download(
  obis_taxa       = "fish",
  regional_poly   = poly,
  obis_outputname = "OBIS_galapagos_fish"
)

# GBIF — supplement, in a single occ_download request.
GBIF_download(
  obis_taxa       = "fish",
  regional_poly   = poly,
  gbif_outputname = "GBIF_galapagos_fish"
)

# Merge OBIS + GBIF (+ any local checklist) into the regional list. A local
# CSV (Genus + Species columns) can live anywhere on disk — no need to copy it
# into the project first.
build_regional_checklist(
  region          = "galapagos",
  taxa            = "fish",
  comb_inputnames = c("OBIS_galapagos_fish", "GBIF_galapagos_fish"),
  local_csvs      = "2016Aug24_Tirado-Sanchez_et_al_Galapagos_Pisces_Checklist.csv"
)
```

This writes two files under `local_database_checklist/`: a species-only list
(`..._for_making_localdb.txt`) to hand to your reference-database builder, and
a list that also keeps any genus-level entries (`..._for_LCA.txt`) — the input
to the next step.

---

## `taxonomize_checklist()`

**Why this matters:** the reconciliation works by walking an assignment up the
ranks, so every checklist species has to carry its full lineage. This step also
updates older names to current canonical NCBI taxonomy, so a synonym on *either*
the checklist or your classifier output still matches instead of silently
failing.

**How:** run it on the `_for_LCA` file. The first run builds a local
`taxonomizr` database (~15 min, several GB downloaded); later runs reuse it.

```{r taxonomize, eval = FALSE}
checklist <- taxonomize_checklist(
  input    = "local_database_checklist/comprehensive_galapagos_fish_list_for_LCA.txt",
  sql_path = "/path/to/accessionTaxa.sql"
)
saveRDS(checklist, "local_database_checklist/galapagos_fish_checklist.rds")
```

---

## `run_regatta()`

**Why this matters:** once the checklist is taxonomized, one call turns your raw
classifier file into a corrected, audited taxonomy table — you don't wire the
preprocessing, reconciliation, and reporting steps together yourself.

**How:** point `run_regatta()` at your results file and the taxonomized
checklist. It detects the format (here an obitools `.tab`, recognized by its
`TAXID` column), resolves the taxIDs to lineages, runs the checklist LCA, and
writes the output bundle.

```{r run, eval = FALSE}
run_regatta(
  input     = "MiFish_obitools.tab",                   # obitools .tab (global DB)
  checklist = "local_database_checklist/galapagos_fish_checklist.rds",
  sql_path  = "/path/to/accessionTaxa.sql"
)
#> -> regatta_out/{ reconcile_checklist/*, regatta_summary.csv, run_log.txt }
```

On this dataset, of the 542 obitools-assigned ASVs (out of 1,438 total):

| match rank | ASVs |
|---|---|
| species | 287 |
| genus   | 229 |
| family  | 16 |
| order   | 1 |
| class   | 9 |

The 229 genus-level and 9 class-level results are exactly the off-target calls
REGATTA exists to catch: globally-confident species assignments to fish that
don't occur in the Galapagos, each walked down to the lowest rank the regional
checklist can actually support.

> **Classified the same ASVs against a second, locally-curated database too?**
> REGATTA can reconcile the two classifier runs against each other before the
> checklist step — see `vignette("REGATTA-two-database")`.

---

## `reconcile_checklist()`

**Why this matters:** this is the core step `run_regatta()` calls for you — and
because it's pure base R, you can watch it work right here, with no internet and
no database, on real data bundled with the package. Seeing each ASV's outcome
makes the correction concrete before you run it on your own table.

**How:** pass a taxonomy table and the taxonomized checklist. The demo below
uses the bundled Galapagos checklist and **12 real obitools assignments (10
genera)** chosen to show every outcome; the two-database vignette runs these
*same 12 ASVs* through the optional step so you can compare the methods. You
would normally pass your *whole* classifier table, not a hand-picked slice.

```{r demo}
library(REGATTA)

checklist <- readRDS(system.file(
  "extdata", "galapagos_fish_checklist.rds", package = "REGATTA"))

# 12 real obitools-assigned ASVs from the worked dataset (ASV_id + 7 ranks + pct_id).
tax <- read.csv(system.file(
  "extdata", "mifish_obitools_example.csv", package = "REGATTA"),
  stringsAsFactors = FALSE)

res <- reconcile_checklist(tax, checklist, output_dir = NULL)
```

`$tracking` records *why* each row changed — the input species, the percent
identity, and the rank at which it matched the regional checklist:

```{r demo-tracking}
res$tracking[, c("before_species", "pct_id", "regatta_match_rank",
                 "after_genus", "after_family")]
```

The checklist step has **three outcomes**, and all three appear here:

* **Kept at species** (the call is on the checklist) — *Abudefduf saxatilis*,
  *Prionurus laticlavius*, *Lutjanus kasmira*, *Caranx melampygus*,
  *Mugil curema*, *Etrumeus micropus*.
* **Downgraded to genus** (species off the checklist, genus on) —
  *Abudefduf taurus* → *Abudefduf*, *Prionurus biafraensis* → *Prionurus*,
  *Epinephelus bontoides* → *Epinephelus*, *Cubiceps squamiceps* → *Cubiceps*.
* **Downgraded to family** (species *and* genus off the checklist) —
  *Archosargus rhomboidalis* → **Sparidae**, *Conodon nobilis* → **Haemulidae**.

The two *within-genus splits* show why this beats a flat percent-identity
cutoff. *Abudefduf saxatilis* and *A. taurus* **both matched the global
database at 100% identity**, yet only *saxatilis* is on the Galapagos
checklist, so *taurus* drops to genus — no cutoff could ever separate them.
Likewise *Prionurus laticlavius* (the Galapagos sawtail surgeonfish) stays at
species while *P. biafraensis* — a **West African** congener confidently called
at 98.8% — drops to genus. The global database simply has no reference for some
Galapagos species, so it names the nearest relative it has; REGATTA keeps the
reliable genus and drops the unreliable species.

`$stats` gives the match-rank distribution for this slice:

```{r demo-stats}
res$stats
```

> Six of these ASVs were *kept at species* here — but three of those calls
> change once a second, locally-curated classifier weighs in.
> `vignette("REGATTA-two-database")` runs the same 12 ASVs through that step and
> compares the two methods side by side.

---

## Interpreting the outputs

**Why this matters:** every REGATTA step that produces a taxonomy table hands
back the same three things, so results drop straight into phyloseq or MetabaR
*and* stay fully traceable — you can always show why a call changed.

**How:** read the three elements (and, when writing to disk, the matching three
CSVs per stage):

| Element | Contents |
|---|---|
| `$result`   | The **exchange format**: exactly the id column + 7 lowercase ranks. Drop straight into a phyloseq `tax_table()` or a MetabaR MOTU table. |
| `$tracking` | Per-ASV audit: before/after at each rank, `regatta_match_rank`, and your passthrough columns. |
| `$stats`    | Step-level counts — the match-rank distribution above. |

## Where to go next

* **A different results format?** `run_regatta()` auto-detects vsearch (LCA +
  `--userout`) and Kraken2/BestTaxon-style name tables. For a table whose
  per-ASV id column isn't `ASV_id` (e.g. a sequence `Hash`), pass
  `id_col = "Hash"`; a long per-sample table is collapsed to one row per ASV
  automatically. See `?run_regatta`.
* **Two reference databases?** If you classified the same ASVs against both a
  global and a locally-curated database, REGATTA can reconcile the two
  classifier runs against each other before the checklist step. See
  `vignette("REGATTA-two-database")` and `?reconcile_global_local`.
* **One pass per primer.** Run the pipeline once per taxonomic group your
  primers target. Mixing groups into one megalist lets off-target
  amplifications slip through.
* **Pre-check your taxon names** with `resolve_taxa()` before a long download,
  and see the README troubleshooting section for GBIF/OBIS specifics.
